```{r}
# 1. LOAD LIBRARIES
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, sf, tmap)

# 2. READ THE DATA
# We use st_read to bring the geopackage into an 'sf' object
roman_roads <- st_read("data/itinere_roads.gpkg")

# 3. CRS CHECK & REPROJECTION
# Excellence Tip: Distance calculations in degrees (WGS84) are inaccurate. 
# We must ensure we are using the British National Grid (meters).
if (st_crs(roman_roads)$epsg != 27700) {
  message("Reprojecting to British National Grid...")
  roman_roads <- st_transform(roman_roads, 27700)
}

# 4. DATA EXPLORATION
# Inspect the first few rows to see the attribute table
print(head(roman_roads))

# 5. INITIAL PLOT
# Using tmap for a high-quality, zoomable preview
tmap_mode("view") # Set to interactive mode

tm_shape(roman_roads) +
  tm_lines(col = "blue", lwd = 2) +
  tm_layout(title = "Roman Road Network Exploration")
```
clip to uk


```{r}
library(janitor)
# --- 1. LOAD LOCAL MSOA BOUNDARIES ---
# Using the path you provided
msoas_uk <- st_read("data/msoa/MSOA_2021_EW_BFE_V8.shp") %>%
  st_transform(27700) %>%
  clean_names()

# --- 2. SPATIAL CLIP ---
# We use the spatial subsetting method to keep only the Roman roads 
# that fall within these MSOA boundaries.
roman_roads_uk <- roman_roads[msoas_uk, ]
```


```{r}
tmap_mode("plot")

msoas_simplified_4326 <- msoas_uk %>%
  st_simplify(dTolerance = 25) %>% 
  st_transform(4326)


# --- 3. VISUAL CHECK ---
# Plotting both to ensure they align perfectly.
# Using 'tmap' for a clean overview.
tm_shape(msoas_uk) +
  tm_polygons(col = "beige", border.col = "grey", border.alpha = 0.5) +
tm_shape(roman_roads_uk) +
  tm_lines(col = "darkblue", lwd = 1) +
  tm_layout(main.title = "Study Area: Roman Roads in England & Wales")
```




Now, let's find the intersections

```{r}
# --- 1. FIND INTERSECTIONS ---
intersections_uk <- st_intersection(roman_roads_uk) %>%
  # Filter to only keep the point geometries (where roads cross)
  st_collection_extract("POINT") %>%
  # IMPORTANT: Break any MultiPoints into individual Point rows
  st_cast("POINT")

# --- 2. CALCULATE DEGREE (CENTRALITY) ---
# Now that every point is its own row, the counts will match
roman_nodes_uk <- intersections_uk %>%
  mutate(
    x = st_coordinates(.)[,1],
    y = st_coordinates(.)[,2]
  ) %>%
  group_by(x, y) %>%
  summarise(degree = n(), .groups = 'drop') %>%
  # Filter for actual junctions (3 or more road-ends meeting)
  # Note: st_intersection of lines usually gives degree 2 for a simple crossing
  filter(degree > 0)

# --- 3. VISUAL CHECK ---
# Let's see the 'Hubs' of Roman Britain
tm_shape(roman_roads_uk) +
  tm_lines(col = "grey", alpha = 0.7) +
tm_shape(roman_nodes_uk) +
  tm_dots(col = "degree", size = 0.7, palette = "YlOrRd", title = "Junction Importance")
```

something does not make sense with the degree calc

```{r}
# 1. DISSOLVE AND NODE
# This 'unions' everything into one big object and then breaks it at every intersection
roman_net_clean <- roman_roads_uk %>%
  st_geometry() %>%      # Work with geometry only for speed
  st_union() %>%         # Dissolve everything into one
  st_node() %>%          # Explicitly add nodes at every true intersection
  st_cast("LINESTRING") %>% # Turn the 'blob' back into individual clean lines
  st_as_sf()             # Back to a spatial data frame

# 2. FIND THE NODES AGAIN
# Since the network is now 'noded', the endpoints of the lines ARE the junctions
roman_nodes_uk <- roman_net_clean %>%
  st_coordinates() %>%
  as_tibble() %>%
  # Group by X and Y to find the meeting points
  group_by(X, Y) %>%
  summarise(degree = n(), .groups = 'drop') %>%
  # FILTERING LOGIC:
  # Degree 1 = Dead end
  # Degree 2 = A simple bend or continuous road (not a junction)
  # Degree 3+ = A true junction (T-junction, Crossroad, etc.)
  filter(degree >= 3) %>%
  st_as_sf(coords = c("X", "Y"), crs = 27700)

# 3. CHECK THE STATS
print("--- Reality Check: Cleaned Degrees ---")
table(roman_nodes_uk$degree)
```


```{r}
tmap_mode("view")

msoas_outline_4326 <- msoas_simplified_4326 |>
  st_union() |>
  st_boundary()


tm_shape(msoas_outline_4326) +
  tm_polygons(fill = "#f8f8f8", col = "white", col_alpha = 0.5) +
tm_shape(roman_roads_uk) +
  tm_lines(col = "gray40", lwd = 0.5, col_alpha = 0.3) +
tm_shape(roman_nodes_uk) +
  tm_dots(
    col = "degree",            # Color the dots based on the 'degree' column
    size = 0.6,           # Also size them by 'degree'
    palette = "viridis",       # Use a vibrant color palette (e.g., viridis, plasma, RdYlBu)
    style = "cat",             # Treat 'degree' as categorical for distinct colors
    title = "Junction Degree"
  ) +
  tm_layout(main.title = "Roman Infrastructure Network", frame = FALSE)
```


Some basic barplot

```{r}
# 1. BASIC SUMMARY STATS
# This gives you the Min, Max, Mean, and Median degree
print("--- Summary Statistics for Roman Junction Degrees ---")
summary(roman_nodes_uk$degree)

# 2. FREQUENCY TABLE
# This tells you exactly how many "Hubs" of each type you have
# e.g., How many 4-way intersections exist in England & Wales?
print("--- Frequency Table (Degree Counts) ---")
degree_counts <- table(roman_nodes_uk$degree)
print(degree_counts)

# 3. VISUALIZATION (For your EDA Chapter)
# A bar chart is the best way to show the distribution of discrete values like 'degree'
library(ggplot2)

ggplot(roman_nodes_uk, aes(x = as.factor(degree))) +
  geom_bar(fill = "darkred", alpha = 0.8) +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5, size=3.5) +
  labs(title = "Frequency of Roman Road Junction Degrees",
       subtitle = "England & Wales Network Analysis",
       x = "Degree (Number of Intersecting Roads)",
       y = "Number of Junctions") +
  theme_minimal()
```



# Calculating roman centrality score - naive

```{r}
# 1. Calculate the distances (Meters)
dist_matrix <- st_distance(st_centroid(msoas_uk), roman_nodes_uk)

# 2. Extract weights (Junction Degrees)
w <- roman_nodes_uk$degree

# 3. Apply the Inverse Square Formula
# Ensure we treat dist_matrix as numeric
dist_num <- as.numeric(dist_matrix)

# This creates the matrix of 3.3 million values
# We wrap it in matrix() to ensure it keeps the right shape for rowSums
gravity_matrix_naive_square_decay <- matrix(1 / ((dist_num + 40)^2), 
                         nrow = nrow(msoas_uk)) %*% diag(w)

# 4. SUM THE ROWS (This is the crucial step)
# This turns the 3,377,760 values into exactly 7,264 values
msoas_simplified_4326$roman_score_naive_square_decay <- rowSums(gravity_matrix_naive_square_decay)

# 5. VERIFY THE DIMENSIONS
# This should now return 7264
length(msoas_simplified_4326$roman_score_naive_square_decay)

```


viz

```{r}
# 1. INTERACTIVE VIEW
tmap_mode("view")

# 2. THE "EQUAL-COUNT" MAP
# Using 'style = "quantile"' ensures that the "lower group" is spread across 
# different colors, showing the subtle differences in connectivity.
tm_shape(msoas_simplified_4326) +
  tm_polygons(
    fill = "roman_score_naive_square_decay",
    palette = "viridis",
    style = "jenks",      # This is the key: forces equal distribution of colors
    n = 10,                  # More bins to see more detail
    title = "Connectivity Deciles",
    alpha = 0.6,
    popup.vars = c("Connectivity" = "roman_score_naive_square_decay"),
    id = "msoa21nm"
  ) +
  tm_layout(main.title = "Roman Connectivity: Distribution by Quantile")

```


# Decay Winning Model

without the root


```{r}
# 1. CALCULATE RAW EXPONENTIAL SIGNALS (No Root)
# dist_km_matrix and w should still be in your environment

# 2. CONVERT TO MATRIX & STRIP UNITS WITHOUT FLATTENING
# We use units::drop_units() or simply matrix(as.numeric()) to keep the shape
dist_km_matrix <- matrix(as.numeric(dist_matrix) / 1000, 
                         nrow = nrow(msoas_uk), 
                         ncol = nrow(roman_nodes_uk))


exp_decay_matrix <- exp(-0.09 * dist_km_matrix)

# Multiply each signal by junction degree
influence_matrix_raw <- sweep(exp_decay_matrix, 2, w, "*")

# 2. SUM WITHOUT THE ROOT
msoas_uk$adj_decay_score <- rowSums(influence_matrix_raw)

# 3. LOG FOR REGRESSION
msoas_uk$log_adj_decay_score <- log(msoas_uk$adj_decay_score + 1)

# 4. CHECK THE RANK-SIZE SKEW
plot_data_re <- msoas_uk %>%
  st_drop_geometry() %>%
  arrange(log_adj_decay_score) %>%
  mutate(rank = row_number())

ggplot(plot_data_re, aes(x = rank, y = log_adj_decay_score)) +
  geom_col(width = 1, fill = "firebrick4") +
  labs(title = "Roman Centrality Score by MSOA",
       subtitle =  "Sorted results, logged",
       x = "MSOA Rank", y = "Log Score") +
  theme_minimal()
```

```{r}

# 2. RENDER THE INTERACTIVE MAP
tmap_mode("plot")

msoas_simplified_4326$log_adj_decay_score = msoas_uk$log_adj_decay_score

tm_shape(msoas_simplified_4326) +
  tm_polygons(fill = "log_adj_decay_score",
              style = "jenks", # Use Jenks to handle the higher skew
              n = 10,
              palette = "YlOrBr",
              alpha = 0.6,
              lwd = 0,
              title = "Roman Centrality Score (Log scale)",
              id = "msoa21nm") +
  
  # Overlay the Roman Roads for context
  tm_shape(roman_roads_uk) +
  tm_lines(col = "black", 
           lwd = 0.7, 
           alpha = 0.7) +
  
  tm_basemap(server = "CartoDB.Positron") +
  tm_view(set.view = c(-1.17, 52.35, 7))
```


```{r}
library(ggplot2)

# 1. GENERATE THE HISTOGRAM
ggplot(msoas_uk, aes(x = log_adj_decay_score)) +
  geom_histogram(fill = "firebrick4", color = "white", bins = 50) +
  # Add a density line to see the 'shape' clearly
  geom_density(aes(y = ..count.. * 0.5), color = "black", size = 1) + 
  labs(title = "Model RE: Raw Exponential (No Root)",
       subtitle = "Distribution of scores across all 7,264 MSOAs",
       x = "Log Connectivity Score", 
       y = "Number of Neighborhoods (Count)") +
  theme_minimal()
```



# Adding Income

```{r}
library(readxl)
library(dplyr)


# 1. Load only the 'Total annual income' sheet
# Replace 'your_income_file.xlsx' with your actual filename
income_raw <- read_excel("data/income.xlsx", 
                         sheet = "Total annual income", 
                         skip = 3)

# 2. Clean the data for the join
# We rename the columns to be 'R-friendly'
income_clean <- income_raw %>%
  select(msoa_code = 'MSOA code', 
         total_income = 'Total annual income (£)')
  # Ensure the income is treated as a number
```


```{r}
# 3. Join it to your Model RE spatial data
# This creates your final dataset: 'msoas_final'
msoas_with_income <- msoas_uk %>%
  left_join(income_clean, by = c("msoa21cd" = "msoa_code")) %>%
  filter(!is.na(total_income)) # Removes any MSOAs without income data
```

```{r}
# Check the number of rows
nrow(msoas_with_income)

# Check the first few rows to ensure income and log_raw_exp_roman are present
head(msoas_with_income %>% select(msoa21cd, log_adj_decay_score, total_income))
```


```{r}
# Transfer the columns to the simplified spatial object
# We use a left_join to ensure we keep the simplified geometry
# Create a simplified version while staying in BNG (27700)
msoas_with_income_simplified <- msoas_with_income %>%
  st_simplify(dTolerance = 25) # Keeps the shape but removes unnecessary detail
```

Plotting side by side maps

```{r}
# Set to plot mode
tmap_mode("plot")

# Map 1: Roman Legacy
map_roman <- tm_shape(msoas_with_income_simplified) +
  tm_polygons(fill = "log_adj_decay_score", 
              style = "jenks", 
              n = 10, 
              palette = "YlOrBr", 
              title = "Roman Score",
              border.alpha = 0) +
  tm_layout(frame = FALSE, 
            main.title = "A: Roman Centrality Score", 
            main.title.size = 0.8,
            # Move legend OUTSIDE the map
            legend.outside = TRUE,
            legend.outside.position = "right",
            # Keep it compact
            legend.title.size = 0.5,
            legend.text.size = 0.4)

# Map 2: Modern Income
map_income <- tm_shape(msoas_with_income_simplified) +
  tm_polygons(fill = "total_income", 
              style = "jenks", 
              n = 10, 
              palette = "Purples", 
              title = "Income (£)",
              border.alpha = 0) +
  tm_layout(frame = FALSE, 
            main.title = "B: Modern Income", 
            main.title.size = 0.8,
            # Move legend OUTSIDE the map
            legend.outside = TRUE,
            legend.outside.position = "right",
            # Keep it compact
            legend.title.size = 0.5,
            legend.text.size = 0.4)

# Arrange side-by-side
tmap_arrange(map_roman, map_income, ncol = 2)
```

second try - more production ready

```{r}
library(tmap)
library(sf)


# 1. SETUP: Titles and Palettes
roman_palette <- "YlOrRd"  # Gold-Orange-Red (Classy for "History")
income_palette <- "BuPu"   # Blue-Purple (Standard for "Wealth")

# 1. VISUAL FIX: Create the "England Only" Road Network
# We use st_intersection to cut the roads exactly at the border of your data.
# (This acts like a cookie cutter)

# 1. Filter for England only (Census codes start with 'E')
# Check your column name! It is usually 'MSOA21CD' or 'msoa11cd'
england_msoas <- msoas_uk %>%
  filter(grepl("^E", msoa21cd)) 

# 2. Unify them into one single outline
# st_union dissolves internal boundaries to create the "Country Shape"
england_outline <- st_union(england_msoas)

# 3. Perform the cut (Your original code)
roman_roads_england <- st_intersection(roman_roads_uk, england_outline)
```


```{r}
# 2. MAP A: THE ROMAN INFRASTRUCTURE
map_roman <- tm_shape(msoas_with_income_simplified) +
  tm_polygons(col = "log_adj_decay_score", 
              title = "Centrality Score (Log scale)", 
              style = "jenks", 
              n = 7, 
              palette = roman_palette, 
              border.alpha = 0,
              colorNA = "grey95") +
  
  # Overlay the Road Network
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", lwd = 0.8, alpha = 0.7) +
  
  # Layout (No Scale Bar to prevent crash)
  tm_layout(main.title = "The Roman Centrality Score", 
            main.title.size = 1.0,
            main.title.fontface = "bold",
            frame = TRUE,
            legend.position = c("left", "top"),
            legend.bg.color = "white",
            legend.bg.alpha = 0.8,
            legend.title.size = 0.8,
            legend.text.size = 0.7)

# 3. MAP B: THE MODERN ECONOMY
map_income <- tm_shape(msoas_with_income_simplified) +
  tm_polygons(col = "total_income", 
              title = "Avg. Income (£)", 
              style = "jenks", 
              n = 7, 
              palette = income_palette, 
              border.alpha = 0,
              colorNA = "grey95") +
  
  # Overlay the Road Network
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", lwd = 0.8, alpha = 0.7) +
  
  # Layout
  tm_layout(main.title = "Average Annual Income (2023)", 
            main.title.size = 1.0,
            main.title.fontface = "bold",
            frame = TRUE,
            legend.position = c("left", "top"),
            legend.bg.color = "white",
            legend.bg.alpha = 0.8,
            legend.title.size = 0.8,
            legend.text.size = 0.7)

# 4. RENDER SIDE-BY-SIDE
# 1. Create the arranged map object
final_comparison <- tmap_arrange(map_roman, map_income, ncol = 2)

# 2. Save it
tmap_save(tm = final_comparison, 
          filename = "imgs/Roman_vs_Income_Comparison.png", 
          width = 10,       # Wider because it's side-by-side (2 maps)
          height = 9,       # Tall enough for England
          units = "in", 
          dpi = 600)

```


## First regression Model

### But first some vis
```{r}
library(ggplot2)
library(patchwork) # For combining plots

# 1. Histogram: Is it a Bell Curve?
p1 <- ggplot(msoas_with_income, aes(x = total_income)) +
  geom_histogram(fill = "purple", color = "white", bins = 50) +
  geom_vline(aes(xintercept = mean(total_income, na.rm = TRUE)), 
             color = "black", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Income",
       subtitle = "Dashed line = Mean Income",
       x = "Total Annual Income (£)", y = "Count") +
  theme_minimal()

# 2. Boxplot: Are there massive outliers?
p2 <- ggplot(msoas_with_income, aes(x = total_income, y = "")) +
  geom_boxplot(fill = "plum", outlier.colour = "red", outlier.shape = 1) +
  labs(title = " Outlier Check (Boxplot)",
       x = "Total Annual Income (£)", y = "") +
  theme_minimal()

# Combine them using patchwork
p1 / p2
```


## Running the Legacy Model

```{r}
# 1. Transform the Income Variable (Log-Log Prep)
msoas_model_ready <- msoas_with_income %>%
  mutate(log_income = log(total_income)) %>%
  filter(!is.na(log_income) & !is.na(log_adj_decay_score))

# 2. RUN THE THESIS MODEL
# Formula: Log(Modern Wealth) ~ Log(Roman Connectivity)
legacy_model <- lm(log_income ~ log_adj_decay_score, data = msoas_model_ready)

# 3. GET THE NUMBERS
summary(legacy_model)
```

```{r}
library(ggplot2)

# Ensure you use the exact dataset and variables from your successful model
ggplot(msoas_model_ready, aes(x = log_adj_decay_score, y = log_income)) +
  # 1. The Data Cloud
  geom_point(alpha = 0.15, color = "darkslategrey", size = 0.8) + 
  
  # 2. The Regression Line
  geom_smooth(method = "lm", color = "firebrick", linewidth = 1, se = TRUE) +
  
  # 3. Victory Labels
  labs(title = "The Roman Gradient (Final Model)",
       subtitle = paste("R-Squared: 0.279 | Coefficient: 0.15 | n = 7,264"),
       x = "Roman Connectivity Score (Log + 1)",
       y = "Modern Household Income (Log £)") +
  
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```


some more viz
```{r}
library(ggplot2)
library(dplyr)

# 1. Create 5 "Tiers" of Roman Connectivity (Quintiles)
viz_data <- msoas_model_ready %>%
  mutate(roman_tier = ntile(log_adj_decay_score, 5)) %>%
  mutate(roman_label = factor(roman_tier, 
                              labels = c("1. Remote", "2. Low", "3. Medium", "4. High", "5. Roman Hub")))

# 2. Plot the Income Distribution for each Tier
ggplot(viz_data, aes(x = roman_label, y = total_income, fill = roman_label)) +
  # The Boxplot shows the median and range
  geom_boxplot(outlier.alpha = 0.1, outlier.color = "gray") +
  
  # Add a trend line connecting the medians
  stat_summary(fun = median, geom = "line", aes(group = 1), color = "firebrick", size = 1) +
  
  scale_fill_brewer(palette = "Oranges") +
  labs(title = "The Roman Premium: Income by Connectivity Tier",
       subtitle = "Median income rises consistently as you move closer to Roman networks",
       x = "Roman Connectivity Tier (Quintiles)",
       y = "Annual Household Income (£)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Mapping the residuals

```{r}

# 1. Extract residuals and add them to your model data frame
# Make sure 'msoas_model_ready' is the object you used to build 'legacy_model'
msoas_model_ready$residuals <- residuals(legacy_model)

# 2. Check to see if it's there
colnames(msoas_model_ready) # You should now see "residuals" at the end


# 1. Ensure tmap is in view mode
tmap_mode("view")



# 2. Build the multi-layered map
# Layer 1: The Residuals (The "Error" in our historical model)
tm_shape(msoas_model_ready %>% st_simplify(dTolerance = 100) # Keeps the shape but removes unnecessary detail
         ) +
  tm_polygons("residuals", 
              palette = "viridis", 
              title = "Model Residuals",
              alpha = 0.5, 
              id = "msoa21nm",
              popup.vars = c("Residual" = "residuals")) +
  
# Layer 2: The Roman Roads (The "Skeleton" of the country)
tm_shape(roman_roads_uk) +
  tm_lines(col = "black", 
           lwd = 1, 
           alpha = 0.6, 
           title.col = "Roman Road Network") +

# Layout and Interactivity
tm_view()
```


```{r}
# 1. Load the necessary libraries
library(readxl)
library(janitor)
library(dplyr)

# 2. Load the 'scores' sheet
# We point to the file path and specify the sheet name
iod_raw <- read_excel("data/iod25.xlsx", sheet = "scores")

# 3. Normalize the column names
# clean_names() converts everything to snake_case (e.g., "Income Score" -> "income_score")
# and handles special characters like "(rate)" or "%" automatically.
iod_scores <- iod_raw %>%
  clean_names()

# 4. Quick check of the new names
colnames(iod_scores)
```

```{r}
library(readxl)
library(readr)
library(janitor)

# 1. Read the Lookup CSV and print columns
lsoa_lookup <- read_csv("data/lsoa_to_msoa.csv") %>% clean_names()
print("--- LSOA to MSOA Lookup Columns ---")
print(colnames(lsoa_lookup))

# 2. Read the Population Excel and print columns
# Specifying the 'pop_denom' sheet as you requested
lsoa_pop <- read_excel("data/lsoa_population.xlsx", sheet = "pop_denom") %>% clean_names()
print("--- Population Data Columns ---")
print(colnames(lsoa_pop))
```


```{r}
# 1. Filter Lookup: Keep only the bridge columns
lsoa_lookup <- lsoa_lookup %>%
  select(lsoa21cd, msoa21cd)

# 2. Filter Population: Keep the code and the 2022 weight
lsoa_pop <- lsoa_pop %>%
  select(lsoa_code_2021, total_population_mid_2022)

# 3. Create the master object by joining your IoD scores to the lookup
# We use inner_join to ensure we "skip" any LSOA without a match
iod_complete <- iod_scores %>%
  inner_join(lsoa_lookup, by = c("lsoa_code_2021" = "lsoa21cd"))

# 4. Finish the master object by joining the population weights
# We overwrite iod_complete to keep it as the single source of truth
iod_complete <- iod_complete %>%
  inner_join(lsoa_pop, by = "lsoa_code_2021")

# 5. Reality Check: Ensure the object is ready for aggregation
# We need to see 'msoa21cd' and 'total_population_mid_2022'
colnames(iod_complete)
```



```{r}
# We perform the aggregation on your current object (iod_complete)
msoa_iod_final <- iod_complete %>%
  group_by(msoa21cd) %>%
  summarise(across(contains("score"), 
                   ~ sum(.x * total_population_mid_2022, na.rm = TRUE) / 
                     sum(total_population_mid_2022, na.rm = TRUE)))

# Reality Check: Check the row count
# It should match your msoas_model_ready row count (approx 7264)
nrow(msoa_iod_final)

# Check the first few rows to ensure the scores look like realistic averages
head(msoa_iod_final)
```

# K Means

Preparing the data
```{r}
library(sf)
library(dplyr)

library(sf)
library(dplyr)

# 1. Calculate the area of each MSOA in your map
# We use st_area() which returns meters squared, then convert to km2
msoa_density_lookup <- msoas_model_ready %>%
  select(msoa21cd, geometry) %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1000000) %>%
  st_drop_geometry() # We only need the numbers for now

# 2. Get the Total Population per MSOA from your IOD/LSOA data
# We sum up the LSOA populations to the MSOA level
msoa_pop_totals <- iod_complete %>%
  group_by(msoa21cd) %>%
  summarise(total_pop = sum(total_population_mid_2022, na.rm = TRUE))

# 3. Join them and calculate Density
# Result: One row per MSOA with a density value
msoa_density_lookup <- msoa_density_lookup %>%
  inner_join(msoa_pop_totals, by = "msoa21cd") %>%
  mutate(pop_density = total_pop / area_km2)

# Verify the result
head(msoa_density_lookup)
```


```{r}
# Join the density we just made with the deprivation scores we aggregated earlier
urban_clustering_input <- msoa_density_lookup %>%
  inner_join(msoa_iod_final, by = "msoa21cd")

# Filter down to exactly the 5 columns you requested
urban_clustering_input <- urban_clustering_input %>%
  select(
    msoa21cd, 
    pop_density, 
    geographical_barriers_sub_domain_score, 
    outdoors_sub_domain_score, 
    wider_barriers_sub_domain_score
  )

# Verify the count and the column names
print(paste("Total rows:", nrow(urban_clustering_input)))
colnames(urban_clustering_input)


```


running the model

```{r}
# 1. Scale the 4 numeric variables (excluding the ID)
scaled_vars <- urban_clustering_input %>%
  select(-msoa21cd) %>%
  scale()

# 2. Run K-means with 4 centers
set.seed(42)
km_result <- kmeans(scaled_vars, centers = 4, nstart = 25)

# 3. Add the cluster ID back to your input dataframe
urban_clustering_input$urbanity_cluster <- as.factor(km_result$cluster)
```



```{r}
library(tidyr)
library(ggplot2)


# 1. Pivot the data to "Long" format for plotting
plot_data <- urban_clustering_input %>%
  pivot_longer(cols = c(pop_density, contains("score")), 
               names_to = "variable", 
               values_to = "value")

# 2. Create the Boxplot
ggplot(plot_data, aes(x = urbanity_cluster, y = value, fill = urbanity_cluster)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~variable, scales = "free_y") +
  theme_minimal() +
  labs(title = "Urbanity Cluster Profiles",
       subtitle = "Comparing Density and IoD Scores across K-means groups",
       x = "Cluster ID",
       y = "Value")
```


### Radar plot
normalization to prepare
```{r}

library(fmsb)
library(dplyr)

# 1. Calculate the mean Z-score for each variable per cluster
# These are our "Centroids"
df_centroids <- urban_clustering_input %>%
  # We use the raw numbers but calculate the Z-scores (Scale)
  # to match your class logic exactly
  mutate(across(c(pop_density, contains("score")), scale)) %>%
  group_by(urbanity_cluster) %>%
  summarise(across(c(pop_density, contains("score")), mean))

# 2. Format for fmsb: it needs a dataframe with specific rows
radar_df <- as.data.frame(df_centroids[,-1])
rownames(radar_df) <- paste("Cluster", df_centroids$urbanity_cluster)

# 3. Add Min/Max rows (Since Z-scores usually fall between -3 and 3, 
# we set these as the chart boundaries)
radar_df <- rbind(rep(3, 4), rep(-3, 4), radar_df)
```

generate

```{r}
# Define colors
colors_border <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3")
colors_fill <- scales::alpha(colors_border, 0.2)

# Set margins
par(mar=c(2, 2, 2, 2))

# Create the Plot
radarchart(radar_df, 
           axistype = 1,
           # Customize the polygon
           pcol = colors_border, pfcol = colors_fill, plwd = 3, plty = 1,
           # Customize the grid
           cglcol = "grey", cglty = 1, axislabcol = "grey", 
           caxislabels = seq(-3, 3, 1.5), cglwd = 0.8,
           # Labels
           vlcex = 0.5, 
           title = "Cluster Centroid Comparison (Z-Scores)")

# Add a legend
legend(x=0.9, y=1.0, legend = rownames(radar_df)[-c(1,2)], 
       bty = "n", pch=20, col=colors_border, cex=1, pt.cex=2)
```

```{r}
# Create a sample of 20 MSOAs per cluster for qualitative analysis
cluster_sample <- urban_clustering_input %>%
  group_by(urbanity_cluster) %>%
  slice_sample(n = 20)

# View the sample in a nice table format
View(cluster_sample)
```



try model

```{r}
library(dplyr)
library(sf)

# 1. Start with your main model object
final_regression_data <- msoas_model_ready %>%
  # Drop geometry to make it a standard dataframe for regression
  st_drop_geometry() %>%
  select(msoa21cd, log_income, log_adj_decay_score) %>%
  
  # 2. Join the Density (Agglomeration)
  left_join(msoa_density_lookup %>% select(msoa21cd, pop_density), by = "msoa21cd") %>%
  
  # 3. Join the "Deep Determinants" (Education & Health)
  # These are the "other variables" that prove it's not just "smart people moving to cities"
  left_join(msoa_iod_final %>% 
              select(msoa21cd, 
                     education_score = education_skills_and_training_score, 
                     health_score = health_deprivation_and_disability_score,
                     employment_score = employment_score_rate), # Optional but good
            by = "msoa21cd") %>%
  na.omit()

# Check what we have
head(final_regression_data)
```


```{r}
# The "Distinction" Model
model_controls <- lm(log_income ~ log_adj_decay_score + 
                       log(pop_density) +   # Control for Density (Agglomeration)
                       education_score +    # Control for Skills (Human Capital)
                       health_score,        # Control for Health
                     data = final_regression_data)

summary(model_controls)
```


Compare noramalized Beta


```{r}
# Install this package if you don't have it
install.packages("lm.beta")
library(lm.beta)

# Calculate the Standardized coefficients
std_model <- lm.beta(model_controls)

# Print the coefficients
print(summary(std_model)$coefficients)
```




Mapping the residuals

```{r}
library(tmap)
library(dplyr)
library(sf)

# 1. Add residuals to the data used in the model
# We use the 'final_regression_data' object because it matches the model rows exactly
final_regression_data$model_residuals <- residuals(model_controls)

# 2. Join this back to the Map Geometry
# We take the ID and the Residual from the data, and join it to the map
map_residuals <- msoas_model_ready %>%
  select(msoa21cd) %>% # Start with just the map shapes
  inner_join(final_regression_data %>% select(msoa21cd, model_residuals), 
             by = "msoa21cd")

# 3. Create the Map
tmap_mode("plot")

tm_shape(map_residuals) +
  tm_polygons("model_residuals",
              title = "Model Residuals (Log Income)",
              palette = "RdBu",      # Red-Blue diverging palette
              midpoint = 0,          # 0 is the center (Perfect prediction)
              style = "cont",        # Continuous gradient
              border.alpha = 0,      # Remove borders for cleaner look
              id = "msoa21cd") +
  tm_layout(main.title = "Where is the model wrong?")
```



```{r}
library(dplyr)
library(spdep)

# 1. Create a "Clean Map" that matches your regression data exactly
# We keep only the MSOAs that are present in 'final_regression_data'
map_clean <- msoas_model_ready %>%
  filter(msoa21cd %in% final_regression_data$msoa21cd)

# 2. Re-build the Neighbours & Weights for this clean map
nb_clean <- poly2nb(map_clean, queen = TRUE)
lw_clean <- nb2listw(nb_clean, style = "W", zero.policy = TRUE)

# 3. NOW run the Moran's I Test
# We add 'zero.policy = TRUE' to handle any islands safely
moran_test <- lm.morantest(model_controls, lw_clean, zero.policy = TRUE)
print(moran_test)

# 4. Run the Lagrange Multiplier Diagnostics
# This checks if you need Lag vs Error
lm_diagnostics <- lm.LMtests(model_controls, lw_clean, test = "all", zero.policy = TRUE)
summary(lm_diagnostics)
```



```{r}
library(spdep)
library(dplyr)

# 1. Filter the Map to match the Regression Data exactly
# We keep only the MSOAs that are present in your 'final_regression_data'
map_clean <- msoas_model_ready %>%
  filter(msoa21cd %in% final_regression_data$msoa21cd)

# 2. Re-build the Neighbours & Weights for this clean map
nb_clean <- poly2nb(map_clean, queen = TRUE)
lw_clean <- nb2listw(nb_clean, style = "W", zero.policy = TRUE)

# 3. NOW run the Moran's I Test (Verification Step 1)
# This checks if the errors are clustered
moran_test <- lm.morantest(model_controls, lw_clean)
print(moran_test)

# 4. Run the Lagrange Multiplier Diagnostics (Verification Step 2)
# This tells us if we need a Spatial Error or Spatial Lag model
lm_diagnostics <- lm.LMtests(model_controls, lw_clean, test = "all")
summary(lm_diagnostics)
```

the spatial error model

```{r}

library(spatialreg)
library(dplyr)

# 1. Safety Step: Ensure Data matches the Map order perfectly
# We use 'map_clean' because that is what 'lw_clean' was built from
final_regression_data_sorted <- final_regression_data %>%
  arrange(factor(msoa21cd, levels = map_clean$msoa21cd))

# 2. Run the Spatial Error Model (SEM)
# 'errorsarlm' is used because your LM tests pointed to "Error"
final_spatial_model <- errorsarlm(log_income ~ log_adj_decay_score +
                                    log(pop_density) +
                                    education_score +
                                    health_score,
                                  data = final_regression_data_sorted,
                                  listw = lw_clean,
                                  zero.policy = TRUE) # Crucial: Ignores the "Island" (Region 6531)

# 3. Reveal the Final Results
summary(final_spatial_model)
```




verify the new results

```{r}
library(tmap)
library(dplyr)
library(sf)

# 1. Extract the "clean" residuals from the spatial model
# The model object has them stored as 'residuals'
final_regression_data_sorted$spatial_residuals <- residuals(final_spatial_model)

# 2. Join back to the map
# We use 'map_clean' which we know matches the data order
map_spatial_check <- map_clean %>%
  select(msoa21cd) %>%
  mutate(spatial_residuals = final_regression_data_sorted$spatial_residuals)

# 3. Plot the Map
tmap_mode("plot")

tm_shape(map_spatial_check) +
  tm_polygons("spatial_residuals",
              title = "Spatial Model Residuals",
              palette = "RdBu",      
              midpoint = 0,          
              style = "cont",      
              border.alpha = 0,
              id = "msoa21cd") +
  tm_layout(main.title = "Did we fix the clustering?")
```

Let's check with Moran's I

```{r}
library(spdep)

# 1. Extract the residuals from the spatial model
# The model object contains them, but we need to make sure they match the weights
spatial_residuals <- residuals(final_spatial_model)

# 2. Run Moran's I Test
# We use the same weights list (lw_clean) as before
final_moran_test <- moran.test(spatial_residuals, lw_clean, zero.policy = TRUE)

# 3. Print the result
print(final_moran_test)
```



# NEW MODEL

# Trying some different model

```{r}
library(sf)
library(dplyr)

# 1. Load the Modern Road Network
# We assume this shapefile contains the relevant strategic roads (Motorways/A-Roads)
modern_roads <- st_read("data/major_road_networks/Major_Road_Network_2018_Open_Roads.shp")

# 2. Check and Fix Coordinate Systems (CRITICAL)
# Both maps must be in British National Grid (EPSG: 27700) to measure meters correctly
if (st_crs(modern_roads)$epsg != 27700) {
  modern_roads <- st_transform(modern_roads, 27700)
}

# Ensure your MSOA map is also 27700 (it likely is, but good to be safe)
msoas_uk_projected <- st_transform(msoas_uk, 27700)

# 3. Calculate Centroids of your MSOAs
# We measure from the "center" of the town
msoa_centroids <- st_centroid(msoas_uk_projected)


# 4. Calculate Distance (The Fast Method)
# Step A: Find the index of the nearest road for each town
nearest_road_index <- st_nearest_feature(msoa_centroids, modern_roads)

# Step B: Extract those specific road segments
nearest_road_segments <- modern_roads[nearest_road_index, ]

# Step C: Calculate the distance only between the town and its specific nearest road
# 'by_element = TRUE' calculates row-by-row, saving massive amounts of RAM
dist_meters <- st_distance(msoa_centroids, nearest_road_segments, by_element = TRUE)

# 5. Add to your dataset and convert to Kilometers
msoas_uk$dist_to_motorway_km <- as.numeric(dist_meters) / 1000

# 6. Log Transform it for the model (Gravity variables are usually logged)
# We add +1 to avoid log(0) if a centroid falls exactly on a road
msoas_uk$log_dist_motorway <- log(msoas_uk$dist_to_motorway_km + 1)

# Check the result
summary(msoas_uk$dist_to_motorway_km)
```

```{r}
library(tmap)

# 1. Ensure the data is in the right object
# (Assuming you ran the previous block and 'msoas_uk' has the new column)

# 2. Set to Interactive Mode
tmap_mode("plot")

# 3. Create the Map
tm_shape(msoas_uk) +
  tm_polygons("dist_to_motorway_km",
              title = "Distance to Major Roads (km)",
              style = "jenks",      # 'Fisher' breaks are good for skewed distance data
              palette = "-magma",    # Dark colors = Far away (Rural), Light = Close (Connected)
              border.alpha = 0,      # Remove borders to see the pattern clearly
              id = "msoa21nm",       # Shows MSOA Name on hover (if available)
              popup.vars = c("Distance (km)" = "dist_to_motorway_km")) +
  tm_layout(main.title = "Modern Connectivity: Distance to Strategic Network")
```
```{r}
library(ggplot2)
library(tmap)
library(sf)

# --- 1. HISTOGRAM & SKEWNESS CHECK ---
# We use ggplot to see the "Long Tail" of rural areas
p1 <- ggplot(msoas_uk, aes(x = dist_to_motorway_km)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "Histogram: Distance to Major Roads",
       subtitle = "Notice the heavy right skew (Long tail of rural areas)",
       x = "Distance (km)", y = "Count of MSOAs") +
  theme_minimal()

print(p1)

# --- 2. THE SKEW PLOT (Density Comparison) ---
# This plot compares the Original (Skewed) vs. Log-Transformed (Normal-ish) data
# This confirms why we need to use 'log()' in the model
p2 <- ggplot(msoas_uk, aes(x = dist_to_motorway_km)) +
  geom_density(fill = "red", alpha = 0.3) +
  labs(title = "Density Plot (The Skew)",
       subtitle = "Most towns are close to roads (peak on left), but some are very far.",
       x = "Distance (km)") +
  theme_minimal()

# Optional: See how Log-transform fixes it
p3 <- ggplot(msoas_uk, aes(x = log(dist_to_motorway_km + 1))) +
  geom_density(fill = "green", alpha = 0.3) +
  labs(title = "Log-Transformed Density",
       subtitle = "This is the variable that will actually go into your model (Bell curve).",
       x = "Log(Distance)") +
  theme_minimal()

print(p2)
print(p3) # You can compare the Red plot (Bad) vs Green plot (Good)

# --- 3. THE MAP WITH ROADS OVERLAY ---
tmap_mode("plot")

tm_shape(msoas_uk) +
  # Layer 1: The MSOAs colored by distance
  tm_polygons("dist_to_motorway_km",
              title = "Dist. to Major Road (km)",
              style = "fisher",
              palette = "-magma",    # Dark = Far, Light = Close
              border.alpha = 0,      # No borders for a cleaner look
              id = "msoa21nm",
              popup.vars = c("Distance (km)" = "dist_to_motorway_km")) +
  
  # Layer 2: The Actual Road Network on top
  tm_shape(modern_roads) + 
  tm_lines(col = "black", 
           lwd = 1.5,                # Line width
           alpha = 0.5) +            # Slightly transparent
  
  tm_layout(main.title = "Verification: Do the light areas follow the black lines?")

```
Seeing the log transform on a map

```{r}
library(tmap)
library(sf)

# 1. Create the Log-Transformed Variable (if you haven't already)
# We add +1 so that 0km becomes log(1) = 0
msoas_uk$log_dist_motorway <- log(msoas_uk$dist_to_motorway_km + 1)

# 2. Map it with the Road Overlay
tmap_mode("plot")

tm_shape(msoas_uk %>% st_simplify(dTolerance = 25)) +
  # Layer 1: The Log-Transformed Distance
  tm_polygons("log_dist_motorway",
              title = "Log Distance to Network",
              style = "jenks",        # Continuous gradient (smooth colors)
              palette = "-viridis",  # Dark = Rural (High Log Dist), Yellow = Connected (Low Log Dist)
              border.alpha = 0,
              id = "msoa21nm",
              popup.vars = c("Log Score" = "log_dist_motorway", 
                             "Real Dist (km)" = "dist_to_motorway_km")) +
  
  # Layer 2: The Actual Road Network (Verification)
  tm_shape(modern_roads) + 
  tm_lines(col = "white",           # White lines stand out better against Viridis
           lwd = 0.8, 
           alpha = 0.2) +
  
  tm_layout(main.title = "Visual Check: Log-Distance vs. Real Roads")

```



# Distance to Charing Cross (London)
```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(tmap)

# --- 1. CALCULATE DISTANCE TO LONDON ---

# Define London (Charing Cross) in British National Grid (EPSG: 27700)
london_point <- st_point(c(530034, 180381)) %>% 
  st_sfc(crs = 27700) %>% 
  st_as_sf()

# Calculate Distance from MSOA centroids
# Ensure msoas_uk is in the correct CRS
msoas_uk_projected <- st_transform(msoas_uk, 27700)
msoa_centroids <- st_centroid(msoas_uk_projected)

# Calculate distance in meters, convert to km
msoas_uk$dist_to_london_km <- as.numeric(st_distance(msoa_centroids, london_point)) / 1000

# Create the Log-Transformed variable (The one for the model)
msoas_uk$log_dist_london <- log(msoas_uk$dist_to_london_km + 0.0001)
```


```{r}
# --- 2. CHECK SKEW (HISTOGRAM & DENSITY) ---

# Plot A: Histogram
p1_lon <- ggplot(msoas_uk, aes(x = dist_to_london_km)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Histogram: Distance to London",
       subtitle = "Notice the spread: London is 0, Scotland is >600km",
       x = "Distance (km)", y = "Count") +
  theme_minimal()

# Plot B: Density Comparison (Raw vs Log)
p2_lon <- ggplot(msoas_uk, aes(x = dist_to_london_km)) +
  geom_density(fill = "red", alpha = 0.3) +
  labs(title = "Raw Density (Skewed)", x = "Distance (km)") +
  theme_minimal()

p3_lon <- ggplot(msoas_uk, aes(x = log_dist_london)) +
  geom_density(fill = "green", alpha = 0.3) +
  labs(title = "Log-Transformed Density (Normal-ish)", 
       subtitle = "This shape is much better for regression.",
       x = "Log(Distance)") +
  theme_minimal()

# Display plots
print(p1_lon)
print(p2_lon)
print(p3_lon)
```


```{r}
# --- 3. THE MAP (YOUR STYLE) ---

tmap_mode("plot") # Or "view" if you want interactive

tm_shape(msoas_uk %>% st_simplify(dTolerance = 25)) +
  # Layer 1: The Log-Transformed Distance to London
  tm_polygons("log_dist_london",
              title = "Log Dist to London",
              style = "jenks",        # Continuous gradient
              palette = "-viridis",  # Yellow = Close (London), Purple = Far
              border.alpha = 0,
              id = "msoa21nm",
              popup.vars = c("Log Score" = "log_dist_london", 
                             "Real Dist (km)" = "dist_to_london_km")) +
  
  # Layer 2: Modern Roads Overlay (Context)
  # Keeping this faint helps show scale/geography even on the London map
  tm_shape(modern_roads) + 
  tm_lines(col = "white", 
           lwd = 0.8, 
           alpha = 0.2) +
  
  # Layer 3: Add a star for London so you can see the "Epicenter"
  tm_shape(london_point) +
  tm_dots(col = "red", size = 0.5, shape = 21, title = "London") +
  
  tm_layout(main.title = "Visual Check: The 'Gravity' of London")
```


the three way map check
```{r}
library(tmap)

# Set mode to plot (Static is better for side-by-side comparison than interactive)
tmap_mode("plot")

# --- MAP 1: DISTANCE TO LONDON (Gravity) ---
# The "Ripples" map we just made
m1 <- tm_shape(msoas_uk %>% st_simplify(dTolerance = 100)) +
  tm_polygons("log_dist_london",
              title = "1. Gravity (London)",
              style = "jenks",       # Jenks shows the 'ripples' best
              n = 5,
              palette = "-viridis",
              border.alpha = 0,
              legend.show = FALSE) + # Hide legend to save space
  tm_shape(london_point) +
  tm_dots(col = "red", size = 0.5) + 
  tm_layout(frame = FALSE, title = "Dist. to London")

# --- MAP 2: ROMAN INDEX (History) ---
# Overlaying the actual Roman Roads (White lines)
m2 <- tm_shape(msoas_uk %>% st_simplify(dTolerance = 100)) +
  tm_polygons("log_adj_decay_score",
              title = "2. Roman Index",
              style = "cont",        # Continuous for density
              palette = "magma",     # Light = High Roman Density
              border.alpha = 0,
              legend.show = FALSE) +
  tm_shape(roman_roads) +            # <--- YOUR ROMAN ROAD SHAPEFILE HERE
  tm_lines(col = "white", lwd = 0.5, alpha = 0.5) +
  tm_layout(frame = FALSE, title = "Roman Network")

# --- MAP 3: DISTANCE TO MOTORWAY (Modern) ---
# Overlaying the Modern Roads (Black lines for contrast or White)
m3 <- tm_shape(msoas_uk %>% st_simplify(dTolerance = 100)) +
  tm_polygons("log_dist_motorway",
              title = "3. Modern Access",
              style = "cont",
              palette = "-viridis",  # Dark = Rural, Light = Connected
              border.alpha = 0,
              legend.show = FALSE) +
  tm_shape(modern_roads) +           # <--- YOUR MODERN ROAD SHAPEFILE
  tm_lines(col = "white", lwd = 0.5, alpha = 0.3) +
  tm_layout(frame = FALSE, title = "Modern Network")

# --- ARRANGE THEM SIDE BY SIDE ---
tmap_arrange(m1, m2, m3, ncol = 3)
```



check some correlation

```{r}
library(dplyr)
library(sf)

# 1. Join the new variables to your final regression dataset
# We join by MSOA Code (msoa21cd) to ensure the rows match perfectly
final_data_with_dist <- final_regression_data_sorted %>%
  left_join(msoas_uk %>% st_drop_geometry() %>% select(msoa21cd, log_dist_london, log_dist_motorway), 
            by = "msoa21cd")

# 2. Select just the 3 contenders
correlation_data <- final_data_with_dist %>%
  st_drop_geometry() %>%
  select(log_adj_decay_score,    # The Hero (Roman)
         log_dist_london,        # The Rival 1 (Gravity)
         log_dist_motorway)      # The Rival 2 (Modern Access)

# 3. Run the Test
cor_matrix <- cor(correlation_data, use = "complete.obs")

# Print it nicely
print(round(cor_matrix, 2))
```



```{r}
library(janitor)
library(dplyr)
library(readr)

# 1. Load the data
# Adjust the path if necessary
hours_data <- read_csv("data/census2021-hours-worked-msoa.csv")

# 2. Clean the column names
hours_data_clean <- hours_data %>% 
  clean_names()

# 3. Print the cleaned names
print(colnames(hours_data_clean))
```


```{r}
library(janitor)
library(dplyr)
library(spatialreg)

# 1. LOAD & CLEAN
hours_raw <- read_csv("data/census2021-hours-worked-msoa.csv", show_col_types = FALSE) %>%
  clean_names()

# 2. CREATE THE VARIABLE (Percentage working 49+ hours)
hours_clean <- hours_raw %>%
  mutate(
    # Calculate % (Multiply by 100 for readability, or keep as 0-1 decimal)
    # We use 0-1 decimal to match other vars usually, but let's stick to what fits your data.
    # Let's use 0-100% to make the coefficient easier to read.
    pct_long_hours = (hours_worked_full_time_49_or_more_hours_worked / 
                      hours_worked_total_all_usual_residents_aged_16_years_and_over_in_employment_the_week_before_the_census) * 100
  ) %>%
  # Select only what we need to join
  select(msoa21cd = geography_code, pct_long_hours)

# 3. JOIN TO YOUR MAIN DATA (The "Robust" Dataset)
# We join this to the dataset that already has the Distance variables
final_data_complete <- final_data_with_dist %>%
  left_join(hours_clean, by = "msoa21cd")

```


```{r}
head(final_data_complete)
```


```{r}
library(ggplot2)
library(tmap)
library(sf)
library(dplyr)

# 1. Join the new variable to the Map Object (msoas_uk)
# We assume 'hours_clean' was created in the previous step
msoas_map_data <- msoas_uk %>%
  left_join(hours_clean, by = c("msoa21cd" = "msoa21cd"))

# --- VISUAL 1: HISTOGRAM ---
# Check the spread. Is it a Bell Curve? Or skewed?
p1_work <- ggplot(msoas_map_data, aes(x = pct_long_hours)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "Histogram: % Working Long Hours (49+)",
       subtitle = "Most places have 10-15% overwork. Where are the 30%+ areas?",
       x = "Percentage (%)", y = "Count of MSOAs") +
  theme_minimal()

# --- VISUAL 2: DENSITY (SKEW CHECK) ---
# Percentages usually don't need Log-transforming, but let's check.
p2_work <- ggplot(msoas_map_data, aes(x = pct_long_hours)) +
  geom_density(fill = "orange", alpha = 0.5) +
  labs(title = "Density Plot: Is the data skewed?",
       subtitle = "If this looks roughly like a bell curve, we DON'T need to log it.",
       x = "Percentage (%)") +
  theme_minimal()

print(p1_work)
print(p2_work)

# --- VISUAL 3: THE MAP ---
# This will show us WHERE the hard workers live.
tmap_mode("view")

tm_shape(msoas_map_data %>% st_simplify(dTolerance = 50)) +
  tm_polygons("pct_long_hours",
              title = "% Working 49+ Hours",
              style = "jenks",
              n = 6,
              palette = "Reds",      # Red = High Overwork (Danger Zone)
              border.alpha = 0,
              alpha = 0.3,
              id = "msoa21nm") +

  
  tm_layout(main.title = "Geography of Overwork")
```


```{r}
library(dplyr)
library(car)      # For VIF
library(janitor)

# 1. FINAL DATA MERGE (Just to be safe)
# We ensure all 3 new variables are in the same place
final_data_boss <- final_regression_data_sorted %>%
  # Join Distance variables
  left_join(msoas_uk %>% st_drop_geometry() %>% select(msoa21cd, log_dist_london, log_dist_motorway), 
            by = "msoa21cd") %>%
  # Join Work Hours variable
  left_join(hours_clean, by = "msoa21cd")

# 2. RUN THE OLS MODEL
boss_model_ols <- lm(log_income ~ log_adj_decay_score +    # The Hero
                                  log(pop_density) +       # Agglomeration
                                  education_score +        # Human Capital
                                  health_score +           # Health
                                  log_dist_motorway +      # Modern Access (Mediator)
                                  log_dist_london +        # Gravity (Confounder)
                                  pct_long_hours,          # Work Ethic
                     data = final_data_boss)

# 3. PRINT THE RESULTS
print("--- THE VERDICT ---")
summary(boss_model_ols)

# 4. CHECK MULTICOLLINEARITY (The Safety Check)
print("--- VIF SCORES (Are variables fighting?) ---")
vif(boss_model_ols)
```

Lagrange on global
```{r}
# 4. Run the Lagrange Multiplier Diagnostics
# This checks if you need Lag vs Error
lm_diagnostics_boss_ols <- lm.LMtests(boss_model_ols, lw_clean, test = "all", zero.policy = TRUE)
summary(lm_diagnostics_boss_ols)
```


spatial err boss

```{r}
library(spatialreg)

# 1. DEFINE THE FORMULA
# (Identical to your OLS, but now fed into the spatial engine)
final_formula <- log_income ~ log_adj_decay_score +    # History
                              log(pop_density) +       # Density
                              education_score +        # Skills
                              health_score +           # Health
                              log_dist_motorway +      # Modern Access
                              log_dist_london +        # Gravity
                              pct_long_hours           # Work Ethic

# 2. RUN THE SPATIAL ERROR MODEL
# Note: This might take 30-60 seconds to run because of the complexity
final_spatial_model <- errorsarlm(final_formula, 
                                  data = final_data_boss, 
                                  listw = lw_clean,    # Your spatial weights matrix
                                  zero.policy = TRUE,
                                  method = "LU")       # Fast calculation method

# 3. PRINT THE RESULTS
summary(final_spatial_model)
```


# GWR MODEL
```{r}
library(spgwr)
library(sf)
library(tmap)

# 1. FIX THE DATA (Re-attach Geometry)
# We join your regression data back to the map shapes to ensure it has geometry
gwr_data_sf <- msoas_uk %>%
  select(msoa21cd) %>%
  left_join(final_data_boss, by = "msoa21cd") %>%
  # Filter out missing data (GWR hates NAs)
  filter(!is.na(log_income), 
         !is.na(log_adj_decay_score),
         !is.na(log_dist_london)) 

# 2. CONVERT TO "SPATIAL" (The format GWR needs)
gwr_data_sp <- as(gwr_data_sf, "Spatial")

# 3. CALCULATE BANDWIDTH (This finds the 'Window Size')
# NOTE: This step might take 2-5 minutes. Don't worry if it pauses.
print("Calculating optimal bandwidth...")
bw_optimal <- gwr.sel(log_income ~ log_adj_decay_score + 
                        log(pop_density) + 
                        education_score + 
                        health_score + 
                        log_dist_motorway + 
                        log_dist_london + 
                        pct_long_hours, 
                      data = gwr_data_sp, 
                      adapt = TRUE)

print(paste("Optimal Bandwidth (Proportion of neighbors):", round(bw_optimal, 4)))

# 4. RUN THE GWR MODEL
print("Running GWR Model...")
gwr_model <- gwr(log_income ~ log_adj_decay_score +
                   log(pop_density) +
                   education_score +
                   health_score +
                   log_dist_motorway +
                   log_dist_london +
                   pct_long_hours,
                 data = gwr_data_sp,
                 adapt = bw_optimal,
                 hatmatrix = TRUE,
                 se.fit = TRUE)

# 5. VISUALIZE THE RESULTS
# Convert results back to SF for mapping
results_sf <- st_as_sf(gwr_model$SDF)

```




```{r}
gwr_results_sf_simplified <-
  results_sf %>% st_simplify(dTolerance = 500)
```


```{r}
# jtt
# 1. Extract Roman Centrality coefficients
roman_coef <- results_sf$log_adj_decay_score

# Summary statistics
summary(roman_coef)
sd(roman_coef)

# 2. Local R²
local_r2 <- results_sf$localR2

# Summary statistics
summary(local_r2)
sd(local_r2)

# 3. Count areas where Roman effect is positive vs negative
table(sign(roman_coef))

# 4. Optional: map the Roman coefficient (for visualization)
library(ggplot2)
ggplot(results_sf) +
  geom_sf(aes(fill = log_adj_decay_score)) +
  scale_fill_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(title = "GWR Roman Centrality Coefficient Across MSOAs")

```


```{r}
library(ggplot2)
library(ggplot2)

ggplot(results_sf, aes(x = "United Kingdom", y = localR2)) +
  # CHANGE 1: Set 'width' to 0.5 (or lower) to make the shape skinnier
  geom_violin(fill = "#4682B4", color = "transparent", alpha = 0.3, width = 0.5) +
  
  # CHANGE 2: Reduce boxplot width slightly to match the new violin size
  geom_boxplot(width = 0.05, fill = "white", color = "black", outlier.shape = NA) +
  
  labs(title = "Local R-Squared Distribution",
       subtitle = "Variation across all MSOAs",
       y = "Adjusted Local R-Squared",
       x = "") +
  
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_blank(), 
    panel.grid.major.x = element_blank(),
    # CHANGE 3: Force the plot to be taller (Portrait mode)
    aspect.ratio = 1.5
  )
```


```{r}
# 1. Check the range to confirm the crazy numbers
summary(results_sf$log_adj_decay_score) # or log_adj_decay_score


library(ggplot2)

# 1. Sort the data from lowest to highest coefficient
results_sf$rank_roman <- rank(results_sf$log_adj_decay_score)

# 2. Create the Index Plot
ggplot(results_sf, aes(x = rank_roman, y = log_adj_decay_score)) +
  # Use points to show the data
  geom_point(color = "steelblue", alpha = 0.5, size = 0.5) +
  
  # Add the Zero line
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  
  # Annotate the "Cliff" (The outliers)
  annotate("text", x = 500, y = -50, label = "Outliers", 
           color = "red", size = 3, fontface = "italic") +
  
  # Labels
  labs(title = "Rank-size Distribution of Roman Effects",
       subtitle = "Ordered coefficients for all MSOAs",
       x = "MSOA Rank (Ordered by Coefficient Size)",
       y = "Coefficient Value") +
  
  theme_minimal()
```


```{r}
# staying with sig woth jtt
library(ggplot2)
library(dplyr)
library(sf)

# 1. Compute t-values for Roman Centrality
results_sf$t_roman <- results_sf$log_adj_decay_score / results_sf$log_adj_decay_score_se

# 2. Filter only significant coefficients (|t| > 1.96)
results_sig <- results_sf %>% filter(abs(t_roman) > 1.96)

# 3. Filter out extreme values using percentiles (1st-99th)
coef_lower <- quantile(results_sig$log_adj_decay_score, 0.05)
coef_upper <- quantile(results_sig$log_adj_decay_score, 0.95)
results_sig_clean <- results_sig %>%
  filter(log_adj_decay_score >= coef_lower & log_adj_decay_score <= coef_upper)

# 4. Summary statistics of filtered & cleaned coefficients
roman_coef_clean <- results_sig_clean$log_adj_decay_score
summary(roman_coef_clean)
sd(roman_coef_clean)
table(sign(roman_coef_clean))  # positive vs negative

# 5. Optional: summary of local R² for cleaned areas
summary(results_sig_clean$localR2)
sd(results_sig_clean$localR2)

# 6. Map the cleaned coefficients
ggplot(results_sig_clean) +
  geom_sf(aes(fill = log_adj_decay_score)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey50") +
  theme_minimal() +
  labs(
    title = "GWR: Significant Roman Centrality Coefficients (Filtered for Extremes)",
    fill = "Roman Coef"
  )



```





```{r}

# Map: Where is the Roman Effect Strongest?
tmap_mode("plot")
tm_shape(gwr_results_sf_simplified) +
  tm_polygons("log_adj_decay_score",
              title = "Local Roman Effect (Coefficient)",
              style = "jenks",      # Highlights the extremes
              palette = "PRGn",
              midpoint = 0,
              alpha = 1,
              border.alpha = 0.02,
              popup.vars = c("Roman Effect" = "log_adj_decay_score", 
                             "Significance (t)" = "log_adj_decay_score_se")) +
  tm_layout(title = "Where History Matters: Local Roman Beta Coefficients")   + 

# Layer 2: The Roman Roads (The Black Skeleton)
tm_shape(roman_roads_uk) +
  tm_lines(col = "black",
           lwd = 0.5,              # Thicker lines to make them stand out
           alpha = 0.3,          # Dark and visible
           popup.vars = c("Road Name" = "Name"))
```

```{r}



# Get the summary of the GWR results
results_summary <- summary(gwr_model$SDF)


  
print("--- COMPARE THE POWER ---")
# 1. Look at Roman Roads (Your Map)
print("Range of Roman Coefficients (Min to Max):")
print(summary(results_sf$log_adj_decay_score))



# 2. Look at Education (The Big Fish)
print("Range of Education Coefficients (Min to Max):")
print(summary(results_sf$education_score))
```


```{r}
# # 1. VISUAL FIX: Create the "England Only" Road Network
# # We use st_intersection to cut the roads exactly at the border of your data.
# # (This acts like a cookie cutter)
# 
# print("Clipping Roman Roads to England boundary...")
# # We unify the MSOAs to create one single "England" shape for cleaner cutting
# england_outline <- st_union(gwr_results_sf_simplified) 
# 
# # Perform the cut
# roman_roads_england <- st_intersection(roman_roads_uk, england_outline)
```


```{r}
library(tmap)

# 1. Define the Breaks to handle the outliers
# Anything below -0.15 or above 0.15 is treated as an "Extreme/Outlier"
# The middle zone (-0.05 to 0.05) is "Weak/Neutral"
my_breaks <- c(-100, -0.15, -0.05, 0, 0.05, 0.15, 100)

# 2. Define Clean Labels for the Legend
my_labels <- c("Outlier (< -15%)", 
               "Strong Negative", 
               "Weak Negative", 
               "Weak Positive", 
               "Strong Positive", 
               "Outlier (> +15%)")

# 3. Plot the Map
tmap_mode("plot")

tm_shape(gwr_results_sf_simplified) +
  tm_polygons("log_adj_decay_score",
              title = "Local Roman Effect\n(Elasticity)",
              style = "fixed",          # CHANGED: Fixed breaks to handle outliers
              breaks = my_breaks,
              labels = my_labels,       # CHANGED: Custom labels
              palette = "PRGn",         # Purple (Neg) to Green (Pos)
              midpoint = 0,
              alpha = 1,
              border.alpha = 0.0,       # No borders for cleaner look
              colorNA = "grey90",       # Missing values are grey
              popup.vars = c("Roman Effect" = "log_adj_decay_score")) +

  # Layer 2: The Roman Roads (The Black Skeleton)
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", 
           lwd = 0.8,                   # Slightly thicker than 0.5 for visibility
           alpha = 0.4) +               # Semi-transparent
  
  # Layout settings
  tm_layout(main.title = "Where History Matters: Local Roman Coefficients",
            main.title.size = 1.1,
            # legend.position = c("right", "top"),
            legend.outside.position = "right",
            legend.outside = TRUE,
            legend.bg.color = "white",
            legend.bg.alpha = 0.8,
            frame = FALSE)
```


or other approach

```{r}
# Check the range of the coefficients (The "Roman Effect")
summary(gwr_results_sf_simplified$log_adj_decay_score)

# Check the t-values (Significance)
# Note: You need to calculate t-values if they aren't there
# t = coefficient / standard_error
gwr_results_sf_simplified$t_value <- gwr_results_sf_simplified$log_adj_decay_score / 
                                     gwr_results_sf_simplified$log_adj_decay_score_se

summary(gwr_results_sf_simplified$t_value)
```


```{r}
# library(tmap)
# library(dplyr)
# 
# 
# 
# library(dplyr)
# library(sf)
# 
# # Step 1: Compute t-values
# sig_map_data <- gwr_results_sf_simplified %>%
#   mutate(
#     t_val = log_adj_decay_score / log_adj_decay_score_se
#   )
# 
# # Step 2: Filter "actually Roman" areas (raw Roman score > 25th percentile)
# threshold <- quantile(gwr_data_sf$log_adj_decay_score, 0.5, na.rm = TRUE)
# sig_map_data <- sig_map_data %>%
#   left_join(
#     st_drop_geometry(gwr_data_sf) %>% select(msoa21cd, raw_roman_score = log_adj_decay_score),
#     by = "msoa21cd"
#   )
# 
# # Step 3: Keep only significant & historically Roman areas
# sig_map_data <- sig_map_data %>%
#   mutate(
#     coef_sig = ifelse(
#       abs(t_val) > 1.96 &        # significant
#       raw_roman_score > threshold, # actually Roman
#       log_adj_decay_score,
#       NA
#     )
#   )
# 
# 
# # sig_map_data <- gwr_results_sf_simplified %>%
# #   mutate(
# #     # Calculate t-stat if missing
# #     t_val = log_adj_decay_score / log_adj_decay_score_se,
# #     
# #     # Create a "Clean" coefficient column: 
# #     # If t-value is between -1.96 and 1.96, set to NA (Grey).
# #     # Also filter out the crazy Cornwall outliers (> 5 or < -5)
# #     coef_sig = ifelse(abs(t_val) > 1.96 & abs(log_adj_decay_score) < 5, 
# #                       log_adj_decay_score, 
# #                       NA)
# #   )
# 
# # 2. PLOT THE "TRUTH MAP"
# tmap_mode("plot")
# 
# tm_shape(sig_map_data %>% st_simplify(dTolerance = 100)) +
#   tm_polygons("coef_sig",
#               title = "Significant Roman Effect\n(Local Coefficient)",
#               style = "fisher",       # "fisher" finds the best natural groups automatically
#               palette = "PRGn",       # Purple (Neg) -> White -> Green (Pos)
#               midpoint = 0,           # Ensures 0 is White
#               colorNA = "grey90",     # Insignificant areas are Grey
#               textNA = "Not Significant",
#               border.alpha = 0,
#               popup.vars = c("Coefficient" = "log_adj_decay_score", "t-value" = "t_val")) +
#   
#   # Add Roads for Context
#   tm_shape(roman_roads_england) +
#   tm_lines(col = "black", lwd = 0.7, alpha = 0.4) +
#   
#   # Layout settings
#   tm_layout(main.title = "Where History Matters: Local Roman Coefficients",
#             main.title.size = 1.5,
#             # legend.position = c("right", "top"),
#             legend.outside.position = "right",
#             legend.outside = TRUE,
#             legend.bg.color = "white",
#             legend.bg.alpha = 0.8,
#             frame = FALSE)
```


```{r}
### 2 side-by-side with london area
library(dplyr)
library(sf)
library(tmap)
library(classInt) # Required for manual breaks

results_sf$msoa21cd <- gwr_data_sf$msoa21cd


# --- DATA PREP (Your Code) ---
# 1. Compute t-values
sig_map_data <- results_sf %>%
  mutate(t_val = log_adj_decay_score / log_adj_decay_score_se)

# 2. Filter "actually Roman" (Thresholding)
threshold <- quantile(gwr_data_sf$log_adj_decay_score, 0.5, na.rm = TRUE)
sig_map_data <- sig_map_data %>%
  left_join(st_drop_geometry(gwr_data_sf) %>% select(msoa21cd, raw_roman_score = log_adj_decay_score),
            by = "msoa21cd")

# 3. Keep only significant & historically Roman areas
sig_map_data <- sig_map_data %>%
  mutate(
    coef_sig = ifelse(
      abs(t_val) > 1.96 & raw_roman_score > threshold, 
      log_adj_decay_score, 
      NA
    )
  )

# --- VISUALIZATION PREP ---

# A. Define Common Breaks (CRITICAL STEP)
# We calculate the breaks *once* on the full dataset so both maps share the scale.
# "fisher" style finds natural clusters in the data.
my_breaks <- classIntervals(sig_map_data$coef_sig[!is.na(sig_map_data$coef_sig)], 
                            n = 5, style = "fisher")$brks

# B. Define a Bounding Box for London
# (Approximate British National Grid coordinates for Greater London)
london_bbox <- st_bbox(c(xmin = 503000, xmax = 561000, 
                         ymin = 155000, ymax = 201000), 
                       crs = st_crs(sig_map_data))

# --- MAP 1: ENGLAND (The Overview) ---
map_england <- tm_shape(sig_map_data %>% st_simplify(dTolerance = 100)) +
  tm_polygons("coef_sig",
              title = "Significant Roman Effect\n(Local Coefficient)",
              breaks = my_breaks,      # <--- USE COMMON BREAKS
              palette = "PRGn",
              midpoint = 0,
              colorNA = "grey92",      # Lighter grey for better contrast
              textNA = "Not Significant",
              border.alpha = 0) +
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", lwd = 0.5, alpha = 0.3) +
  tm_layout(main.title = "A. National View",
            main.title.size = 1.2,
            frame = FALSE,
            legend.position = c("left", "bottom")) # Legend on this map

# --- MAP 2: LONDON ZOOM (The Detail) ---
map_london <- tm_shape(sig_map_data) + # No simplify for London (keep detail)
  tm_polygons("coef_sig",
              breaks = my_breaks,      # <--- USE SAME COMMON BREAKS
              palette = "PRGn",
              midpoint = 0,
              colorNA = "grey92",
              border.alpha = 0.1,      # Slight border to see neighborhoods
              legend.show = FALSE) +   # HIDE LEGEND (Shared with Map 1)
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", lwd = 1, alpha = 0.5) + # Thicker roads for zoom
  tm_layout(main.title = "B. London & The 'Spine'",
            main.title.size = 1.2,
            frame = FALSE) +
  tm_view(bbox = london_bbox) # Force the zoom

# --- COMBINE THEM ---
tmap_mode("plot")
tmap_arrange(map_england, map_london, ncol = 2)
```


another try
```{r}
library(dplyr)
library(sf)
library(tmap)
library(classInt)
library(maptiles) # <--- NEW PACKAGE for stable map tiles
library(terra)    # <--- NEW PACKAGE to handle the image

# --- REPAIR STEP: GET NAMES ---
sig_map_data <- sig_map_data %>%
  left_join(st_drop_geometry(msoas_uk) %>% select(msoa21cd, msoa21nm), 
            by = "msoa21cd")

# --- PREP 1: DEFINE LONDON BOX ---
london_bbox_obj <- st_bbox(c(xmin = 500000, xmax = 563000, 
                             ymin = 150000, ymax = 221000), 
                           crs = st_crs(sig_map_data))

# --- PREP 2: FETCH BACKGROUND (The Fix) ---
# We use 'get_tiles' instead of 'read_osm'.
# provider = "CartoDB.Positron" is the modern name for 'CartoLight'
london_bg <- get_tiles(x = london_bbox_obj, 
                       provider = "CartoDB.Positron", 
                       zoom = 12,    # Zoom 12 is perfect for city view
                       crop = TRUE)

# --- PREP 3: FIX ROADS ---
london_roads_clean <- st_crop(roman_roads_england, london_bbox_obj)

# --- VISUALIZATION ---

# Define Common Breaks
my_breaks <- classIntervals(sig_map_data$coef_sig[!is.na(sig_map_data$coef_sig)], 
                            n = 5, style = "fisher")$brks

# MAP 1: ENGLAND (Overview)
map_england <- tm_shape(sig_map_data %>% st_simplify(dTolerance = 100)) +
  tm_polygons("coef_sig",
              title = "Significant Roman Effect\n(Local Coefficient)",
              breaks = my_breaks,
              palette = "PRGn",
              midpoint = 0,
              colorNA = "grey92",
              textNA = "Not Significant",
              border.alpha = 0) +
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", lwd = 0.5, alpha = 0.3) +
  tm_layout(main.title = "National View",
            main.title.size = 1.1,
            frame = FALSE,
            legend.position = c("left", "bottom"))

# MAP 2: LONDON (With 'maptiles' Background)
map_london <- tm_shape(london_bg) +   # <--- LAYER 1: The Raster Background
  tm_rgb() +                          # <--- Function to plot the image
  
  tm_shape(sig_map_data, bbox = london_bbox_obj) + 
  # LAYER 2: The Data (Semi-transparent)
  tm_polygons("coef_sig",
              breaks = my_breaks,
              palette = "PRGn",
              midpoint = 0,
              colorNA = "grey92",    # Dark grey for insignificant areas
              alpha = 0.6,           # 60% opacity to let the map show through
              border.alpha = 0,
              legend.show = FALSE) +
  
  # LAYER 3: The Roman Roads
  tm_shape(london_roads_clean) +
  tm_lines(col = "black", lwd = 1.5) + # Bold black lines
  
  # Layout
  tm_layout(main.title = "London",
            main.title.size = 1.1,
            frame = FALSE)

# --- COMBINE ---
tmap_mode("plot")
tmap_arrange(map_england, map_london, ncol = 2)

```

all eng VIEW

```{r}
library(dplyr)
library(sf)
library(tmap)
library(classInt)

# --- STEP 1: PREP & SIMPLIFY ---
# We create a specific 'lightweight' version for the national map.
# dTolerance = 200 means details smaller than 200m are smoothed out.
england_simple <- sig_map_data %>%
  st_simplify(dTolerance = 200)

# --- STEP 2: DEFINE COMMON BREAKS ---
# Critical: Use the exact same logic as your local maps so colors match
my_breaks <- classIntervals(sig_map_data$coef_sig[!is.na(sig_map_data$coef_sig)], 
                            n = 5, style = "fisher")$brks

# --- STEP 3: VISUALIZATION ---
tmap_mode("view") # Static mode for high-quality export

tm_shape(england_simple) +
  tm_polygons("coef_sig",
              title = "Significant Roman Effect\n(Local Coefficient)",
              breaks = my_breaks,
              palette = "plasma",         # Purple (Neg) <-> Green (Pos)
              midpoint = 0,             # White is zero
              colorNA = "grey92",       # Light grey for insignificant/missing
              textNA = "Not Significant",
              border.alpha = 0,         # No borders (looks cleaner on national map)
              lwd = 0) +
  
  # Add Roman Roads (Thin and faint for context)
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", lwd = 0.3, alpha = 0.6) +
  
  # Layout settings
  tm_layout(main.title = "Where History Matters: National GWR Roman Beta Coefficients",
            main.title.position = "center",
            main.title.size = 1,
            legend.position = c("left", "bottom"),
            legend.bg.color = "white",
            legend.bg.alpha = 0.8,
            frame = FALSE)
```

Brighton, Leicester
```{r}
library(dplyr)
library(sf)
library(tmap)
library(classInt)
library(maptiles)
library(terra)

# --- 1. PREP & REPAIR ---
if(!"msoa21nm" %in% names(sig_map_data)) {
  sig_map_data <- sig_map_data %>%
    left_join(st_drop_geometry(msoas_uk) %>% select(msoa21cd, msoa21nm), by = "msoa21cd")
}

# Define Viewing Windows (~65km wide)
bbox_brighton <- st_bbox(c(xmin = 498000, xmax = 563000, ymin = 85000,  ymax = 150000), crs = st_crs(sig_map_data))
bbox_leicester <- st_bbox(c(xmin = 425000, xmax = 490000, ymin = 270000, ymax = 335000), crs = st_crs(sig_map_data))

# --- 2. COOKIE CUTTER FIX ---
poly_brighton <- st_as_sfc(bbox_brighton)
poly_leicester <- st_as_sfc(bbox_leicester)

data_brighton <- st_intersection(sig_map_data, poly_brighton)
data_leicester <- st_intersection(sig_map_data, poly_leicester)

roads_brighton <- st_intersection(roman_roads_england, poly_brighton) %>% st_union() %>% st_line_merge()
roads_leicester <- st_intersection(roman_roads_england, poly_leicester) %>% st_union() %>% st_line_merge()

# --- 3. BACKGROUNDS ---
bg_brighton  <- get_tiles(bbox_brighton,  provider = "CartoDB.PositronNoLabels", zoom = 10, crop = TRUE)
bg_leicester <- get_tiles(bbox_leicester, provider = "CartoDB.PositronNoLabels", zoom = 10, crop = TRUE)

# --- 4. LABELS ---
label_df <- data.frame(
  label = c("BRIGHTON", "LEICESTER"),
  x = c(531000, 458700), y = c(104500, 304300), city = c("Brighton", "Leicester")
) %>% st_as_sf(coords = c("x", "y"), crs = st_crs(sig_map_data))

# --- 5. VISUALIZATION ---
my_breaks <- classIntervals(sig_map_data$coef_sig[!is.na(sig_map_data$coef_sig)], n = 5, style = "fisher")$brks

tmap_mode("plot")

# MAP A: BRIGHTON (No Legend)
map_brighton <- tm_shape(bg_brighton) + tm_rgb() +
  tm_shape(data_brighton) +
  tm_polygons("coef_sig", 
              breaks = my_breaks, 
              palette = "plasma", 
              midpoint = NA,
              colorNA = "grey", alpha = 0.7, border.alpha = 0,
              textNA = "Insignificant",
              legend.show = FALSE) +  # <--- HIDE LEGEND
  
  tm_shape(roads_brighton) + tm_lines(col = "black", lwd = 1.2) +
  
  tm_shape(label_df %>% filter(city == "Brighton")) +
  tm_text("label", size = 0.7, fontface = "bold", col = "black", shadow = TRUE, bg.color = "white", bg.alpha = 0.5) +
  
  tm_layout(main.title = "The Commuter Spine (Brighton)", 
            main.title.size = 0.9, 
            frame = FALSE)

# MAP B: LEICESTER (With Legend + Caption)
map_leicester <- tm_shape(bg_leicester) + tm_rgb() +
  tm_shape(data_leicester) +
  tm_polygons("coef_sig", 
              title = "Roman Effect",
              breaks = my_breaks, 
              palette = "plasma", 
              midpoint = NA,
              colorNA = "grey", alpha = 0.7, border.alpha = 0,
              textNA = "Insignificant",
              legend.show = TRUE,         # <--- SHOW LEGEND
              legend.is.portrait = TRUE) + 
  
  tm_shape(roads_leicester) + tm_lines(col = "black", lwd = 1.2) +
  
  tm_shape(label_df %>% filter(city == "Leicester")) +
  tm_text("label", size = 0.7, fontface = "bold", col = "black", shadow = TRUE, bg.color = "white", bg.alpha = 0.5) +
  
  # CAPTION CREDIT
  tm_credits("Black lines = Roman Roads\nSource: Itiner-e Digital Atlas", 
             position = c("right", "bottom"), 
             size = 0.5, 
             bg.color = "white", 
             bg.alpha = 0.6) +

  tm_layout(main.title = "The Midlands (Leicester)", 
            main.title.size = 0.9,
            legend.position = c("left", "bottom"), # <--- INSIDE CORNER
            legend.bg.color = "white",
            legend.bg.alpha = 0.6,
            legend.text.size = 0.5,
            legend.title.size = 0.7,
            frame = FALSE)

# --- 6. COMBINE ---
tmap_arrange(map_brighton, map_leicester, ncol = 2)
```

ALL eng 
```{r}
library(dplyr)
library(sf)
library(tmap)
library(classInt)
library(maptiles)
library(terra)

# --- 1. PREP ---
# Simplify to make it fast
england_simple <- sig_map_data %>% st_simplify(dTolerance = 100)

# Define a TIGHT bounding box around England/Wales
# I've manually tweaked these numbers to cut off unnecessary sea/France
bbox_national <- st_bbox(c(xmin = 130000, xmax = 660000, 
                           ymin = 10000,  ymax = 660000), 
                         crs = st_crs(sig_map_data))

# --- 2. FETCH BACKGROUND ---
# Zoom 7 is good for national
bg_national <- get_tiles(bbox_national, provider = "CartoDB.PositronNoLabels", zoom = 7, crop = TRUE)

# --- 3. VISUALIZATION ---
my_breaks <- classIntervals(sig_map_data$coef_sig[!is.na(sig_map_data$coef_sig)], n = 5, style = "fisher")$brks

tmap_mode("plot")

tm_shape(bg_national) + 
  tm_rgb() +
  
  # DATA LAYER
  tm_shape(england_simple) +
  tm_polygons("coef_sig", 
              title = "Roman Effect",
              breaks = my_breaks, 
              palette = "plasma", 
              midpoint = NA,
              colorNA = "grey",
              textNA = "Insignificant",
              alpha = 0.8,
              border.alpha = 0,
              legend.show = TRUE,
              legend.is.portrait = TRUE) +
  
  # ROAD LAYER
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", lwd = 0.5, alpha = 0.5) +
  
  # CREDITS (Bottom Right)
  tm_credits("Black lines = Roman Roads\nSource: Itiner-e Digital Atlas", 
             position = c("right", "bottom"), 
             size = 0.5,             # Smaller text
             bg.color = "white", 
             bg.alpha = 0.4) +

  # LAYOUT FIXES
  tm_layout(main.title = "Where History Matters: National GWR Roman Beta Coefficients", 
            main.title.size = 1.0,
            main.title.position = "center",
        
            legend.bg.color = "white",
            legend.bg.alpha = 0.7,
            # 3. SHRINK LEGEND
            legend.title.size = 0.8,
            legend.text.size = 0.6,
            frame = TRUE)
```
pro map?
```{r}
# ... (Keep your data prep 1 & 2 the same) ...
# --- 1. FETCH BACKGROUND (UPDATED) ---
# "OpenStreetMap" gives you the classic look with town names.
# "CartoDB.Positron" is a lighter alternative if OSM is too colorful.
bg_national <- get_tiles(bbox_national, provider = "OpenStreetMap", zoom = 7, crop = FALSE)

# --- 2. VISUALIZATION ---
m <- tm_shape(bg_national) + 
  tm_rgb(alpha = 0.7) +  # Fade the map slightly so your data pops out
  
  # DATA LAYER
  tm_shape(england_simple) +
  tm_polygons("coef_sig", 
              title = "Roman Effect",
              breaks = my_breaks, 
              palette = "plasma", 
              midpoint = NA,
              colorNA = "grey", 
              textNA = "Insignificant/dropped",
              alpha = 0.7,      # Lower alpha slightly so town names show through
              border.alpha = 0,
              legend.show = TRUE,
              legend.is.portrait = TRUE) +
  
  # ROAD LAYER
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", lwd = 1.4, alpha = 0.2) +

  # CREDITS
  tm_credits("Black lines = Roman Roads\nSource: Itiner-e Digital Atlas", 
             position = c("right", "bottom"), 
             size = 0.9, bg.color = "white", bg.alpha = 0.6) +

  # LAYOUT
  tm_layout(
    main.title = "Where History Matters: Local Roman Coefficients", 
    main.title.size = 1.5,
    main.title.position = "center",
    main.title.fontface = "bold",
    fontfamily = "sans", 
    
    # Legend Position (Adjust as needed)
    legend.position = c(0.75, 0.95), 
    legend.bg.color = "white",
    legend.bg.alpha = 0.8,
    legend.frame = TRUE,
    legend.title.size = 1,
    legend.text.size = 0.9,
    frame = FALSE,
    # Force camera to look at England (fixes blank map issue)
    bbox = england_simple
  )

# Save
tmap_save(m, filename = "imgs/beta_Map_Pro.png", dpi = 600, height = 10, width = 8)
```


```{r}
library(dplyr)
library(sf)

print("=== 🏛️ RECOVERING IDS & FINDING WINNERS ===")

# --- STEP 0: THE FIX ---
# The GWR output lost the IDs. We copy them back from the original input data.
# Note: Ensure 'gwr_data_sf' is the exact dataframe you fed into the GWR model.
gwr_results_sf_simplified$msoa21cd <- gwr_data_sf$msoa21cd

# (Optional: Check case sensitivity. If your ID is "MSOA21CD" in capitals, use that instead)


# --- STEP 1: JOIN NAMES ---
# Now the left_join will work because 'msoa21cd' exists in both.
results_named <- gwr_results_sf_simplified %>%
  left_join(st_drop_geometry(msoas_uk) %>% select(msoa21cd, msoa21nm), 
            by = "msoa21cd")

# --- STEP 2: DEFINE "ACTUALLY ROMAN" ---
# Calculate the threshold from the INPUT data (the raw Roman scores)
threshold <- quantile(gwr_data_sf$log_adj_decay_score, 0.25) 

# --- STEP 3: JOIN RAW SCORES & CALCULATE T-VALUES ---
clean_assets <- results_named %>%
  st_drop_geometry() %>%
  # Join raw Roman score to filter out "fake" positives (Cornwall effect)
  left_join(st_drop_geometry(gwr_data_sf) %>% 
              select(msoa21cd, raw_roman_score = log_adj_decay_score), 
            by = "msoa21cd") %>%
  # Calculate t-values (Coefficient / Standard Error) to check significance
  # Note: Adjust column names if your GWR output names them differently (e.g. 'log_adj_decay_score_SE')
  mutate(t_val = log_adj_decay_score / log_adj_decay_score_se)

# --- STEP 4: REVEAL THE WINNERS ---
print("=== 🏛️ TRUE ROMAN ASSETS (Significant, Positive, & Historically Roman) ===")

top_winners <- clean_assets %>%
  filter(abs(t_val) > 1.96) %>%           # 1. Statistically Significant
  filter(log_adj_decay_score > 0) %>%     # 2. Positive Economic Impact
  filter(raw_roman_score > threshold) %>% # 3. Actually Near a Road (Removes Cornwall)
  arrange(desc(log_adj_decay_score)) %>%  # Sort by strength of effect
  select(msoa21nm, local_coef = log_adj_decay_score, raw_roman_score, t_val) %>%
  head(50)

top_winners
```

try to auto plot winners and losers
```{r}
library(ggplot2)
library(dplyr)
library(sf)

# --- STEP 1: DATA EXTRACTION ---
# We take the map data, drop the shapefiles, and grab the extremes.
ranked_data <- sig_map_data %>%
  st_drop_geometry() %>%
  filter(!is.na(coef_sig)) %>%
  select(msoa21nm, coef_sig) %>%
  arrange(desc(coef_sig)) # Sort High to Low

# Grab Top 10 (Winners) and Bottom 10 (Losers)
top_winners <- head(ranked_data, 5) %>% mutate(Type = "Winner")
top_losers  <- tail(ranked_data, 5) %>% mutate(Type = "Loser")

# Combine them for plotting
plot_data <- bind_rows(top_winners, top_losers)

# Clean MSOA names (Optional: removes redundant text if needed)
# plot_data$msoa21nm <- gsub(" \\(England\\)", "", plot_data$msoa21nm)

# --- STEP 2: THE VISUALIZATION ---
ggplot(plot_data, aes(x = reorder(msoa21nm, coef_sig), y = coef_sig, fill = coef_sig)) +
  
  # 1. The Bars
  geom_col(width = 0.7) +
  
  # 2. Add Value Labels (White text inside the bar for readability)
  geom_text(aes(label = round(coef_sig, 2), 
                hjust = ifelse(coef_sig > 0, 1.2, -0.2)), # Adjust position based on +/-
            color = "white", size = 3.5, fontface = "bold") +
  
  # 3. Flip Coordinates (Make it horizontal)
  coord_flip() +
  
  # 4. Colors (Matches your 'Plasma' Map)
  scale_fill_viridis_c(option = "plasma", name = "Coefficient") +
  
  # 5. Labels & Captions
  labs(title = "The Roman Divide",
       subtitle = "The neighborhoods with the most extreme coefficients",
       x = NULL, # No X label needed (names are self-explanatory)
       y = "Local Roman Coefficient") +
  
  # 6. Professional Theme
  theme_minimal()
```





```{r}
library(ggplot2)
library(dplyr)
library(sf)

# 1. PREPARE THE DATA
# We extract the coefficients and calculate the t-values
hist_data <- gwr_results_sf_simplified %>%
  st_drop_geometry() %>%
  mutate(
    t_val = log_adj_decay_score / log_adj_decay_score_se,
    
    # Classify each town into 3 buckets
    Category = case_when(
      t_val > 1.96  ~ "Significant Positive (Asset)",
      t_val < -1.96 ~ "Significant Negative (Trap)",
      TRUE          ~ "Insignificant (Noise)"
    )
  ) %>%
  # FILTER OUTLIERS: We zoom in to the "Realistic Economy" zone (-1 to +1)
  # This ignores the crazy Cornwall errors so we can see the real pattern.
  filter(log_adj_decay_score > -2 & log_adj_decay_score < 2)

# 2. PLOT THE HISTOGRAM
ggplot(hist_data, aes(x = log_adj_decay_score, fill = Category)) +
  # Create the bars
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.8) +
  
  # Color them correctly
  scale_fill_manual(values = c("grey80", "forestgreen", "purple")) +
  
  # Add a center line at 0 (Neutral)
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", size = 1) +
  
  # Labels and Titles
  labs(title = "Distribution of the Roman Effect (GWR Coefficients)",
       subtitle = "Green = Significant Assets | Purple = Significant Traps | Grey = Noise",
       x = "Local Coefficient (Elasticity)",
       y = "Number of MSOAs (Towns)") +
  
  theme_minimal() +
  theme(legend.position = "top")
```








R^2 mapped

```{r}
# 1. RETRIEVE GLOBAL OLS METRICS
global_model <- lm(log_income ~ log_adj_decay_score + 
                     log(pop_density) + 
                     education_score + 
                     health_score + 
                     log_dist_london + 
                     pct_long_hours, 
                   data = gwr_data_sf)

print("--- MODEL FIGHT: OLS vs. GWR ---")

# 2. CHECK IF AIC IS AVAILABLE (The Crash Fix)
# We look inside the results to see if AIcc exists
if (!is.null(gwr_model$results$AIcc)) {
  ols_aic <- AIC(global_model)
  gwr_aic <- gwr_model$results$AIcc
  
  print(paste("Global OLS AIC:", round(ols_aic, 2)))
  print(paste("GWR AICc:      ", round(gwr_aic, 2)))
  print(paste("Difference:    ", round(ols_aic - gwr_aic, 2)))
  
  if((ols_aic - gwr_aic) > 3) {
    print("RESULT: VICTORY. GWR is statistically superior.")
  }
} else {
  print("WARNING: AICc is missing (Likely because hatmatrix=FALSE).")
  print("Skipping AIC test. Relying on R-Squared below.")
}

# 3. R-SQUARED COMPARISON (This always works)
# OLS R2
ols_r2 <- summary(global_model)$r.squared

# GWR R2 - We calculate the mean of the local R2 values
# Note: In spgwr, localR2 is stored in the SDF object
gwr_r2_mean <- mean(gwr_model$SDF$localR2, na.rm = TRUE)

print(paste("Global OLS R2: ", round(ols_r2, 3)))
print(paste("GWR Mean R2:   ", round(gwr_r2_mean, 3)))

# 4. MAP THE PERFORMANCE (Local R2)
tmap_mode("plot")

tm_shape(gwr_results_sf_simplified) +
  tm_polygons("localR2",
              title = "Local R-Squared\n(Model Fit)",
              style = "cont",         
              palette = "Blues",      # Dark Blue = High Accuracy
              border.alpha = 0,
              popup.vars = c("Local R2" = "localR2")) +
  
  tm_layout(main.title = "Where is the model smartest?",
            # legend.position = c("right", "bottom"),
            legend.outside.position = "right",
            legend.outside = TRUE,
            frame = FALSE)
```

```{r}
# 1. RE-CALCULATE GLOBAL OLS (Just to be sure we compare apples to apples)
global_model <- lm(log_income ~ log_adj_decay_score + 
                     log(pop_density) + 
                     education_score + 
                     health_score + 
                     log_dist_motorway + 
                     log_dist_london + 
                     pct_long_hours, 
                   data = gwr_data_sf)

# 2. EXTRACT THE METRICS
# OLS Scores
ols_aic <- AIC(global_model)
ols_r2  <- summary(global_model)$r.squared

# GWR Scores (Now available because hatmatrix=TRUE)
gwr_aic <- gwr_model$results$AICc
gwr_r2  <- mean(gwr_model$SDF$localR2)

# 3. PRINT THE RESULTS
print("========================================")
print("      FINAL MODEL COMPARISON            ")
print("========================================")
print(paste("Global OLS AIC:   ", round(ols_aic, 2)))
print(paste("GWR AICc:         ", round(gwr_aic, 2)))
print("----------------------------------------")
print(paste("DIFFERENCE (AIC): ", round(ols_aic - gwr_aic, 2)))
print("========================================")

# 4. THE VERDICT
diff <- ols_aic - gwr_aic
if(diff > 3) {
  print("✅ VICTORY: GWR is statistically superior.")
  print("   (The drop in AICc proves spatial non-stationarity exists.)")
} else {
  print("❌ DEFEAT: GWR did not improve the model enough.")
  print("   (The extra complexity wasn't worth it.)")
}

print(" ")
print(paste("Global OLS R-Squared: ", round(ols_r2, 3)))
print(paste("GWR Average R-Squared:", round(gwr_r2, 3)))
```


comparing the 3 models
```{r}
# 1. EXTRACT AIC SCORES
ols_aic <- AIC(global_model)
sem_aic <- AIC(final_spatial_model)
gwr_aic <- gwr_model$results$AICc

# 2. PRINT THE LEADERBOARD
print("==========================================")
print("      🏆 ULTIMATE MODEL CHAMPIONSHIP      ")
print("==========================================")
print(paste("1. Global OLS AIC:   ", round(ols_aic, 1)))
print(paste("2. Spatial Error SEM:", round(sem_aic, 1)))
print(paste("3. Local GWR AICc:   ", round(gwr_aic, 1)))
print("------------------------------------------")

# Calculate Improvements
imp_sem <- ols_aic - sem_aic
imp_gwr <- sem_aic - gwr_aic

print(paste("Improvement OLS -> SEM: ", round(imp_sem, 1), "(Spatial Clustering Effect)"))
print(paste("Improvement SEM -> GWR: ", round(imp_gwr, 1), "(Spatial Non-Stationarity Effect)"))

print("==========================================")
if(gwr_aic < sem_aic & (sem_aic - gwr_aic) > 3) {
  print("✅ FINAL VERDICT: GWR WINS.")
  print("   (The relationship varies significantly across space.)")
} else if(sem_aic < ols_aic) {
  print("⚠️ FINAL VERDICT: SEM WINS.")
  print("   (Spatial clustering matters, but the Roman effect is uniform.)")
}
print("==========================================")
```


# in the GWR are there some areas with roman as the most important variable?

```{r}
# 1. CALCULATE GLOBAL STANDARD DEVIATIONS (The "Equalizer")
# We measure how "spread out" each variable is across the whole country
sd_roman <- sd(gwr_data_sf$log_adj_decay_score, na.rm=TRUE)
sd_educ  <- sd(gwr_data_sf$education_score, na.rm=TRUE)
sd_health <- sd(gwr_data_sf$health_score, na.rm=TRUE)

print(paste("SD Roman:", round(sd_roman, 2)))
print(paste("SD Education:", round(sd_educ, 2)))

# 2. STANDARDIZE THE GWR COEFFICIENTS
# We multiply the coefficient by the SD to get the "Standardized Beta"
# We take Absolute Value (abs) because we care about STRENGTH, not direction (+/-)
gwr_results_sf_simplified$std_beta_roman  <- abs(gwr_results_sf_simplified$log_adj_decay_score * sd_roman)
gwr_results_sf_simplified$std_beta_educ   <- abs(gwr_results_sf_simplified$education_score * sd_educ)
gwr_results_sf_simplified$std_beta_health <- abs(gwr_results_sf_simplified$health_score * sd_health)

# 3. FIND THE WINNER
# We compare the three standardized scores for every single MSOA
gwr_results_sf_simplified$dominant_driver <- apply(
  st_drop_geometry(gwr_results_sf_simplified[, c("std_beta_roman", "std_beta_educ", "std_beta_health")]), 
  1, # For each row
  function(x) {
    # If Roman is the biggest number, return "History"
    # If Education is biggest, return "Education"
    # etc.
    names(x)[which.max(x)] 
  }
)

# Rename for pretty map
gwr_results_sf_simplified$dominant_driver <- factor(
  gwr_results_sf_simplified$dominant_driver,
  levels = c("std_beta_roman", "std_beta_educ", "std_beta_health"),
  labels = c("History (Roman)", "Education", "Health")
)

# 4. MAP THE WINNERS
tmap_mode("plot")

tm_shape(gwr_results_sf_simplified) +
  tm_polygons("dominant_driver",
              title = "Dominant Driver of Wealth",
              palette = c("red", "orange", "blue"), # red=Roman, Orange=Edu, Blue=Health
              border.alpha = 0,
              alpha = 0.8) +
  
  tm_layout(main.title = "The Battle for Dominance: Which Variable Matters Most?",
            frame = FALSE)
```

model validations
```{r}


library(performance)
library(see)       # Required for the plots
library(car)       # For VIF
library(spdep)     # For Moran's I
library(tmap)      # For Mapping


# 1. VISUAL CHECK (Linearity, Normality, Homoscedasticity)
# This creates the "Dashboard" plot you saw on the website.
# It checks: 
# - Linearity (Residuals vs Fitted)
# - Homoscedasticity (Scale-Location)
# - Normality (QQ Plot)
# - Outliers (Cook's Distance)
check_model(global_model, check = c("linearity", "homogeneity", "qq", "outliers"))
```


```{r}
# 1. CALCULATE VIF (If not already done)
vif_scores <- vif(global_model)

# 2. SETUP MARGINS
# The vector is c(bottom, left, top, right)
# We increase the 'left' margin (2nd number) from the default 4 to 12
par(mar = c(5, 12, 4, 2))

# 3. PLOT
barplot(vif_scores, 
        main = "Assumption 4: Multicollinearity (VIF Scores)", 
        horiz = TRUE,         # Horizontal bars
        col = "steelblue",    # Professional blue
        las = 1,              # Force all text to be horizontal
        cex.names = 0.8,      # Shrink text slightly to ensure it fits
        xlim = c(0, 5))       # Set x-axis limit to 5 (the danger zone)

# 4. ADD REFERENCE LINE
# We add a red dashed line at 5. If bars stay left of this, you win.
abline(v = 5, col = "red", lty = 2, lwd = 2)

# Optional: Add a text label for the threshold
text(x = 5, y = 1, labels = "Threshold (5)", col = "red", pos = 3, cex = 0.8)

# 5. RESET MARGINS (Good practice so future plots don't look weird)
par(mar = c(5, 4, 4, 2) + 0.1)
```

```{r}
print("--- ASSUMPTION 5: SPATIAL INDEPENDENCE ---")

# A. EXTRACT RESIDUALS
# We verify the data length matches before adding
if(nrow(gwr_results_sf_simplified) == length(residuals(global_model))) {
  gwr_results_sf_simplified$ols_residuals <- residuals(global_model)
} else {
  print("Warning: Row counts didn't match perfectly. Using safe merge...")
  gwr_results_sf_simplified$ols_residuals <- NA
  gwr_results_sf_simplified$ols_residuals[as.numeric(names(residuals(global_model)))] <- residuals(global_model)
}

# B. MORAN'S I TEST (The Statistical Proof)
# (Using the 6-nearest neighbors weights we created earlier)
moran_test <- lm.morantest(global_model, lw_clean)
print(moran_test)

# C. MAP THE RESIDUALS (The Visual Proof)
# Andrew uses Red/Blue diverging palettes to show clustering.
tmap_mode("plot")

tm_shape(gwr_results_sf_simplified) +
  tm_polygons("ols_residuals",
              title = "OLS Residuals",
              style = "fisher",     # Fisher breaks are good for finding clusters
              palette = "-RdBu",    # Red = Under-prediction, Blue = Over-prediction
              midpoint = 0,
              border.alpha = 0,
              colorNA = "grey90") +
  tm_layout(main.title = "Assumption 5 Check: Are Errors Clustered?",
            legend.position = c("left", "bottom"),
            frame = FALSE)
```



```{r}
library(spdep)

print("--- ASSUMPTION 5: SPATIAL INDEPENDENCE (Moran's I) ---")

# 1. RUN THE TEST
# H0 (Null Hypothesis): Errors are random (Spatial Independence).
# H1 (Alt Hypothesis): Errors are clustered (Spatial Autocorrelation).
moran_result <- lm.morantest(global_model, lw_clean, alternative = "two.sided")

print(moran_result)

# 2. INTERPRETATION HELPER
if(moran_result$p.value < 0.05) {
  print("❌ RESULT: SIGNIFICANT CLUSTERING (Assumption Failed).")
  print("The p-value is < 0.05, meaning the residuals are NOT random.")
  print("This confirms that OLS is biased and justifies using SEM/GWR.")
} else {
  print("✅ RESULT: RANDOM ERRORS (Assumption Passed).")
}
```


DIAGNOSTICS ERROR SPATIAL MODEL

```{r}
library(spatialreg)
library(ggplot2)
library(gridExtra)

# 1. DEFINE THE DASHBOARD FUNCTION (With Smaller Fonts)
create_spatial_dashboard <- function(resid, fitted, model_name) {
  
  # Define a common theme update to apply to all plots
  # Change 'size = 10' to make it even smaller or larger
  my_theme <- theme_minimal() + 
              theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5),
                    axis.title = element_text(size = 9),
                    axis.text  = element_text(size = 8))

  # A. Linearity (Residuals vs Fitted)
  p1 <- ggplot(data.frame(f = fitted, r = resid), aes(x = f, y = r)) +
    geom_point(alpha = 0.3, color = "blue") +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 0.8) +
    geom_smooth(method = "loess", se = FALSE, color = "darkgreen", size = 0.8) +
    labs(title = paste("1. Linearity:", model_name), 
         x = "Fitted Values", y = "Residuals") +
    my_theme

  # B. Homoscedasticity (Scale-Location)
  std_resid <- sqrt(abs(resid))
  p2 <- ggplot(data.frame(f = fitted, sr = std_resid), aes(x = f, y = sr)) +
    geom_point(alpha = 0.3, color = "purple") +
    geom_smooth(method = "loess", se = FALSE, color = "red", size = 0.8) +
    labs(title = paste("2. Homoscedasticity:", model_name), 
         x = "Fitted Values", y = "√|Standardized Residuals|") +
    my_theme

  # C. Normality (Histogram)
  p3 <- ggplot(data.frame(r = resid), aes(x = r)) +
    geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "skyblue", color = "white") +
    geom_density(color = "darkblue", size = 0.8) +
    labs(title = paste("3. Normality:", model_name), x = "Residuals", y = "Density") +
    my_theme

  # D. Normality (Q-Q Plot)
  p4 <- ggplot(data.frame(r = resid), aes(sample = r)) +
    stat_qq(alpha = 0.3) +
    stat_qq_line(color = "red", size = 0.8) +
    labs(title = "4. Q-Q Plot", x = "Theoretical", y = "Sample") +
    my_theme

  # Combine them into a grid
  grid.arrange(p1, p2, p3, p4, ncol = 2)
}

# 2. RUN FOR SPATIAL ERROR MODEL (SEM)
print("--- GENERATING SEM DASHBOARD ---")
sem_resid  <- residuals(final_spatial_model)
sem_fitted <- fitted(final_spatial_model)

create_spatial_dashboard(sem_resid, sem_fitted, "SEM")

# 3. RUN FOR GWR MODEL
print("--- GENERATING GWR DASHBOARD ---")
# Extract GWR math
gwr_resid  <- gwr_model$SDF$gwr.e
gwr_fitted <- gwr_model$SDF$pred

create_spatial_dashboard(gwr_resid, gwr_fitted, "GWR")

```

```{r}
library(spdep)
library(tmap)
library(sf)

print("=========================================")
print("   ASSUMPTION 5: INDEPENDENCE CHECK      ")
print("=========================================")

# -------------------------------------------------------
# 1. PREPARE THE RESIDUALS
# -------------------------------------------------------
# OLS Residuals
ols_resid <- residuals(global_model)

# SEM Residuals
sem_resid <- residuals(final_spatial_model)

# GWR Residuals (Extracted from the SpatialDataFrame)
gwr_resid <- gwr_model$SDF$gwr.e 

# Add them to your spatial data for mapping
# (We use a safe merge to ensure rows align)
map_data <- gwr_results_sf_simplified
map_data$resid_ols <- ols_resid
map_data$resid_sem <- sem_resid
map_data$resid_gwr <- gwr_resid

# -------------------------------------------------------
# 2. STATISTICAL TEST (Moran's I)
# -------------------------------------------------------
print("--- TEST RESULTS: MORAN'S I ---")

# A. OLS (The Baseline)
# We expect this to FAIL (p < 0.05)
ols_moran <- lm.morantest(global_model, lw_clean)
print(paste("1. OLS Moran's I:", round(ols_moran$estimate[1], 4), 
            "(p-value:", format.pval(ols_moran$p.value, digits=3), ")"))

# B. SEM (The Spatial Fix)
# This usually lowers the clustering significantly.
sem_moran <- moran.test(sem_resid, lw_clean)
print(paste("2. SEM Moran's I:", round(sem_moran$estimate[1], 4), 
            "(p-value:", format.pval(sem_moran$p.value, digits=3), ")"))

# C. GWR (The Local Fix)
# This should be the closest to zero (Random).
gwr_moran <- moran.test(gwr_resid, lw_clean)
print(paste("3. GWR Moran's I:", round(gwr_moran$estimate[1], 4), 
            "(p-value:", format.pval(gwr_moran$p.value, digits=3), ")"))

# --------------------------------
```


```{r}
library(tmap)
library(sf)

# 1. PREPARE THE DATA
# Ensure the residuals are attached to your map object
map_data <- gwr_results_sf_simplified

# Attach SEM Residuals
map_data$resid_sem <- residuals(final_spatial_model)

# Attach GWR Residuals
map_data$resid_gwr <- gwr_model$SDF$gwr.e 

# 2. DEFINE SHARED MAP SETTINGS
# We create a reusable layout to make the comparison fair and clean
my_map_style <- tm_layout(legend.position = c("left", "bottom"), 
                          frame = FALSE, 
                          title.size = 0.9,
                          legend.bg.color = "white",
                          legend.bg.alpha = 0.8)

# 3. CREATE SEM MAP (Spatial Error Model)
map_sem <- tm_shape(map_data) +
  tm_polygons("resid_sem", 
              title = "SEM Residuals", 
              style = "fisher",      # Smart breaks for clustering
              palette = "-RdBu",     # Red = Under-predicted, Blue = Over-predicted
              midpoint = 0, 
              border.alpha = 0,
              colorNA = "grey90") +
  
  # Add the Roman Roads (England Subset)
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", lwd = 0.5, alpha = 0.3) +
  
  tm_layout(title = "1. SEM Residuals") + my_map_style

# 4. CREATE GWR MAP (Local Model)
map_gwr <- tm_shape(map_data) +
  tm_polygons("resid_gwr", 
              title = "GWR Residuals", 
              style = "fisher", 
              palette = "-RdBu", 
              midpoint = 0, 
              border.alpha = 0,
              colorNA = "grey90") +
  
  # Add the Roman Roads (England Subset)
  tm_shape(roman_roads_england) +
  tm_lines(col = "black", lwd = 0.5, alpha = 0.3) +
  
  tm_layout(title = "2. GWR Residuals (Clean)") + my_map_style

# 5. PLOT SIDE-BY-SIDE
tmap_arrange(map_sem, map_gwr, ncol = 2)
```




# Some extra visualizations
```{r}
library(ggplot2)
library(dplyr)
library(sf)

# 1. DATA PREP (Same as before)
target_places <- data.frame(
  msoa21nm = c("City of London 001", "St Albans 014", "Bath and North East Somerset 010", 
               "Chester 006", "York 008", "Dover 004", "Leicester 015", 
               "Newcastle upon Tyne 018", "Carlisle 005", "Brighton and Hove 015", 
               "Manchester 015", "Birmingham 030", "Leeds 025", 
               "King's Lynn and West Norfolk 002", "Cornwall 040"),
  display_label = c("City of London 001 (Londinium)", "St Albans 014 (Verulamium)", "Bath 010 (Aquae Sulis)",
                    "Chester 006 (Deva Victrix)", "York 008 (Eboracum)", "Dover 004 (Dubris)",
                    "Leicester 015 (Ratae)", "Newcastle 018 (Pons Aelius)", "Carlisle 005 (Luguvalium)",
                    "Brighton and Hove 015", "Manchester 015 (Mamucium)", "Birmingham 030",
                    "Leeds 025 (West Yorkshire)", "King's Lynn and West Norfolk 002 (Rural East)", "Cornwall (The Frontier)")
)

plot_data <- msoas_with_income_simplified %>%
  st_drop_geometry() %>%
  inner_join(target_places, by = "msoa21nm") %>%
  select(display_label, log_adj_decay_score) %>%
  arrange(desc(log_adj_decay_score))

# 2. CREATE THE BLUE PLOT
ggplot(plot_data, aes(x = reorder(display_label, log_adj_decay_score), 
                      y = log_adj_decay_score)) +
  
  # A. The Bars: Professional "Steel Blue"
  geom_col(fill = "#4682B4", width = 0.7) + 
  
  # B. Flip to Horizontal
  coord_flip() +
  
  # C. Data Labels: Standard Sans-Serif Font
  geom_text(aes(label = sprintf("%.1f", log_adj_decay_score)), 
            hjust = -0.3, 
            # family = "serif",  <-- REMOVED to revert to normal
            color = "#333333", 
            size = 4,
            fontface = "bold") +
  
  # D. Titles
  labs(title = "The Roman Hierarchy: selected examples",
       subtitle = "Centrality Score (Log) of selected neighborhoods",
       caption = "Source: Itiner-e Roman Roads Dataset & ONS Census 2021",
       x = NULL,
       y = "Roman Centrality Score") +
  
  # E. Theme (Standard Sans-Serif)
  theme_minimal() + 
  theme(
      # 1. Force the title to align to the whole PLOT, not just the inner panel
      plot.title.position = "plot", 
      plot.caption.position = "plot",
  
      # 2. Use hjust = 0 for left alignment
      plot.title = element_text(face = "bold", size = 18, color = "#1a1a1a", hjust = 0),
      plot.subtitle = element_text(size = 12, color = "#4a4a4a", hjust = 0),
      plot.caption = element_text(size = 8, color = "grey60", margin = margin(t = 10), hjust = 0), # Optional: left align caption too?
      
      # ... rest of your theme settings ...
      axis.text.y = element_text(size = 11, color = "#2b2b2b"),
      axis.text.x = element_blank(),     
      panel.grid.major.y = element_blank(), 
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(color = "grey90", linetype = "dotted"),
      plot.margin = margin(20, 20, 20, 20)
    )
  
  # F. Expansion
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```




```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(sf)

# 1. PREPARE THE DATA
# We take the clean dataframe 'gwr_data_sf'
plot_data_controls <- gwr_data_sf %>%
  st_drop_geometry() %>%
  # Create the Log Transformation for Density
  mutate(log_pop_density = log(pop_density)) %>%
  select(education_score, health_score, log_pop_density) %>%
  
  # Rename for clean plot labels
  rename(
    "Education Score" = education_score,
    "Health Score" = health_score,
    "Population Density (Log)" = log_pop_density
  ) %>%
  
  # Reshape to long format for faceting
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")

# 2. CREATE THE DASHBOARD PLOT
ggplot(plot_data_controls, aes(x = Value, fill = Variable)) +
  
  # A. The Histograms
  # bins=30 is standard, 'color="white"' adds the clean separators between bars
  geom_histogram(bins = 40, color = "white", alpha = 0.85) +
  
  # B. Facet Wrap (Side-by-side panels)
  facet_wrap(~Variable, scales = "free", ncol = 3) +
  
  # C. Custom "Journalism" Palette
  scale_fill_manual(values = c("#2A9D8F", "#E9C46A", "#F4A261")) +
  
  # D. Labels & Titles
  labs(title = "The Control Variables",
       subtitle = "Distribution of Education, Health, and Density across England",
       caption = "Source: ONS Census 2021. Density is log-transformed.",
       y = "Count of Neighborhoods",
       x = "Score / Value") +
  
  # E. Theme Styling
  theme_minimal() +
  theme(
    # Typography
    plot.title = element_text(face = "bold", size = 16, hjust = 0, color = "#1a1a1a"),
    plot.subtitle = element_text(size = 11, color = "grey40", margin = margin(b = 20)),
    strip.text = element_text(face = "bold", size = 11, color = "#333333"),
    
    # Legend (Not needed since titles explain it)
    legend.position = "none",
    
    # Clean Grid
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    
    # Margins
    plot.margin = margin(20, 20, 20, 20)
  )
```



```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(sf)

# 1. PREPARE THE DATA
# We select all relevant variables from the final dataset
plot_data_full <- gwr_data_sf %>%
  st_drop_geometry() %>%
  
  # Ensure we have the log versions where needed for the plot
  # (Adjust column names if yours differ slightly, e.g., 'dist_london' vs 'log_dist_london')
  mutate(
    log_pop_density = log(pop_density),
    # If your data doesn't have these exact names, standardise them here:
    # log_dist_london = log(dist_london), 
    # log_dist_motorway = log(dist_motorway)
  ) %>%
  
  # Select the 8 key variables
  select(
    log_income,          # The Outcome
    log_adj_decay_score,   # The Main Predictor
    education_score,       # Control 1
    health_score,          # Control 2
    log_pop_density,       # Control 3
    log_dist_london,       # Control 4
    log_dist_motorway,     # Control 5
    pct_long_hours         # Control 6
  ) %>%
  
  # Rename for professional plot labels
  rename(
    "Annual Avg Income (Log £)" = log_income,
    "Roman Centrality (Log)" = log_adj_decay_score,
    "Education deprivation" = education_score,
    "Health deprivation" = health_score,
    "Population Density (Log)" = log_pop_density,
    "Dist to London (Log)" = log_dist_london,
    "Dist to Motorway (Log)" = log_dist_motorway,
    "% Long Hours" = pct_long_hours
  ) %>%
  
  # Reshape for Faceting
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")

# 2. CREATE THE FULL DASHBOARD
ggplot(plot_data_full, aes(x = Value, fill = Variable)) +
  
  # A. The Histograms
  # bins=40 gives good detail. Color=white separates the bars.
  geom_histogram(bins = 40, color = "white", linewidth = 0.1, alpha = 0.9) +
  
  # B. Facet Wrap
  # ncol=4 creates a nice 2x4 grid
  facet_wrap(~Variable, scales = "free", ncol = 4) +
  
  # C. Color Palette
  # We use 'viridis' (discrete) to give each variable a distinct, professional color
  scale_fill_viridis_d(option = "plasma", direction = -1) +
  
  # D. Labels
  labs(title = "The Model Ingredients",
       subtitle = "Distributions of all variables used in the final analysis",
       caption = "Source: ONS Census 2021, Itiner-e, 2025 Index Of Deprivation",
       y = "Neighborhood Count",
       x = "Value") +
  
  # E. Theme
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0, color = "#1a1a1a"),
    plot.subtitle = element_text(size = 12, color = "grey40", margin = margin(b = 20)),
    
    # Strip text (Panel titles)
    strip.text = element_text(face = "italic", size = 8, color = "#333333"),
    
    # Remove Legend (Redundant)
    legend.position = "none",
    
    # Clean up the grid
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 7),
    axis.text.x = element_text(size = 7),
    
    # Margins
    plot.margin = margin(20, 20, 20, 20)
  )
```




global model vis

```{r}
library(ggplot2)
library(dplyr)
library(tibble)

# 1. CREATE DATA FRAME FROM YOUR RESULTS
# We manually input the coefficients you pasted
model_results <- tribble(
  ~Variable,              ~Estimate,   ~Type,
  "Health Deprivation",   -0.1334,     "Control",
  "Dist to London",       -0.0908,     "Control",
  "Dist to Motorway",     -0.0163,     "Control",
  "Pop Density",           0.0099,     "Control",
  "Roman Centrality",      0.0082,     "Focus",    # <--- The Tiny Hero
  "Education Deprivation",-0.0046,     "Control"
)

# 2. CREATE THE PLOT
ggplot(model_results, aes(x = reorder(Variable, abs(Estimate)), y = Estimate, fill = Type)) +
  
  # A. The Bars
  geom_col(width = 0.7) +
  
  # B. Flip coordinates (so names are on the left)
  coord_flip() +
  
  # C. Colors: Make Roman distinct ("Roman Red" vs "Grey Controls")
  scale_fill_manual(values = c("grey70", "#8B0000")) +
  
  # D. Add the Numbers (Percent Impact) for the reader
  geom_text(aes(label = sprintf("%.1f%%", Estimate * 100), 
                hjust = ifelse(Estimate > 0, -0.2, 1.2)), # Adjust position based on +/-
            size = 4, fontface = "bold", color = "#333333") +
  
  # E. Labels
  labs(title = "The False Verdict: 'History Doesn't Matter'",
       subtitle = "Estimated impact of variables on Income (Global OLS Model)",
       caption = "Note: Coefficients represent % change in income for a unit increase in log predictor.",
       y = "Impact on Income (Coefficient)",
       x = NULL) +
  
  # F. Theme
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    legend.position = "none",
    axis.text.y = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_blank()
  )
```




Caterppillar?

```{r}

library(ggplot2)
library(dplyr)
library(sf)

# --- STEP 1: PREP DATA ---
# 1. Drop geometry (we don't need the map shapes for a scatter plot)
# 2. Filter for only significant values (remove NAs)
# 3. Sort them from lowest to highest
# 4. Add a "Rank" column (1, 2, 3... N) for the X-axis
plot_data <- sig_map_data %>%
  st_drop_geometry() %>%
  filter(!is.na(coef_sig)) %>%
  arrange(coef_sig) %>%
  mutate(rank = row_number())

# --- STEP 2: PLOT ---
ggplot(plot_data, aes(x = rank, y = coef_sig, color = coef_sig)) +
  
  # 1. The Points (One dot per MSOA)
  geom_point(size = 0.5, alpha = 0.8) +
  
  # 2. Zero Line (The "Neutral" Horizon)
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30", size = 0.8) +
  
  # 3. Annotation Text (Optional but helpful)
  annotate("text", x = 500, y = min(plot_data$coef_sig), 
           label = "The 'Trap'", color = "grey30", fontface = "bold", hjust = 0) +
  annotate("text", x = max(plot_data$rank) - 500, y = max(plot_data$coef_sig), 
           label = "The 'Spine'", color = "grey30", fontface = "bold", hjust = 1) +

  # 4. Colors (Matching your map's Plasma theme)
  scale_color_viridis_c(option = "plasma", name = "Coefficient") +
  
  # 5. Labels
  labs(title = "The Spectrum of Roman Influence",
       subtitle = paste0("Sorted coefficients for ", nrow(plot_data), " significant MSOAs"),
       x = "Neighborhood Rank (Sorted Low to High)",
       y = "Roman Effect (GWR Coefficient)") +
  
  # 6. Theme
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```
roman roads standalone

```{r}
# 1. Save the map object to a variable first
my_final_map <- tm_shape(bg_roman) + 
  tm_rgb() +
  tm_shape(roman_roads_england) +
  tm_lines(col = "#1a1a1a", lwd = 1.5, alpha = 0.85) +
  tm_compass(type = "8star", position = c("right", "top"), size = 2, text.color = "grey30") +
  tm_scale_bar(breaks = c(0, 50, 100), position = c("right", "bottom"), text.size = 0.6, color.dark = "grey30") +
  tm_credits("Source: Itiner-e Digital Atlas", position = c("right", "bottom"), size = 0.5, col = "grey50") +
  tm_layout(main.title = "The Roman Road Network in England",
            main.title.position = "center", main.title.fontface = "bold", main.title.size = 1.3,
            # CRITICAL: Set margins to 0 to fill the frame
            inner.margins = c(0, 0, 0, 0),
            outer.margins = c(0, 0, 0, 0),
            frame = TRUE, frame.lwd = 1.5, bg.color = "white")

# 2. Save it with a TALL aspect ratio (e.g., Height is double the Width)
tmap_save(tm = my_final_map, 
          filename = "imgs/Roman_Roads_England_BIG.png", 
          width = 5,         # Narrow width
          height = 9,        # Tall height (Portrait orientation)
          units = "in",      # Inches
          dpi = 600)         # High resolution for print
```

